### 2021-08-22 draft.Rmd
initial commit

---
title: "데이터 분석가의 R 패키지 개발 도전기"
author: "이영록"
date: "2021/11/19 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: '16:9'      
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

class: inverse, left, middle

1. 데이터 분석 문서화에서 패키지 개발로 전환
  - 배경
  - 작업 과정
2. 데이터 분석가에게 유용한 개발자 관점
  - 프로젝트 관리
  - R 코딩 습관
  - 메타 프로그래밍
  - 단위 테스트

---
class: inverse, center, middle

# 데이터 분석 문서화에서 패키지 개발로 전환

## 배경

---

# 데이터마이닝 기법과 응용

```{r book, echo = FALSE, fig.cap="데이터미아닝 기법과 응용 / 한나래출판사 / 전치혁 저", out.height='400px', fig.align='center'}
knitr::include_graphics("images/book-cover.jpg")
```

---

# `{bookdown}`을 이용한 R 예제 책 만들기

```{r online-book-webshot, include=FALSE, cache=TRUE}
webshot::webshot(
  url = "https://youngroklee-ml.github.io/data-mining-book/index.html",
  file = "images/onlinebook-main.jpg",
  zoom = 3
)
```

```{r online-book, echo = FALSE, fig.cap="데이터마이닝 with R / Published with bookdown / 전치혁, 이혜선, 이종석, 이영록 공저", out.height='400px', fig.align='center'}
knitr::include_graphics("images/onlinebook-main.jpg")
```

---

# 교재용 R 패키지 작성의 필요성

--

- 알고리즘의 단계별 결과를 교재 내용대로 보여주기 위해 함수를 세분화할 필요

--

- 기존 R 패키지가 교재에 설명된 알고리즘과 정확히 일치하지 않는 부분에 대한 혼동을 줄일 필요

--

- 사용된 기존 R 패키지의 버전업에 따른 교재의 R 스크립트 수정을 줄일 필요

---
class: inverse, center, middle

# 데이터 분석 문서화에서 패키지 개발로 전환

## 작업 과정

---

# 현실적인 scope 정의

- 참고교재
- 실무용 아님
- 제한된 유연성

---

# 편리한 작업환경 선택

- RStudio IDE
- `{devtools}`, `{usethis}`, `{testthat}`, etc.
- Git and GitHub

```{r r-packages-book, echo = FALSE, fig.cap="R Packages / O'Reilly / Hadley Wickham", out.width='25%', fig.align='center'}
knitr::include_graphics("images/r-pkgs-cover.png")
```

---

# 문서 내 R 스크립트 분류

- 데이터 생성 --> data-raw/
- 함수 생성 --> R/
- 데이터 분석  --> vignettes/

---

# Hello World

Install the **xaringan** package from [Github](https://github.com/yihui/xaringan):

```{r eval=FALSE, tidy=FALSE}
remotes::install_github("yihui/xaringan")
```

--

You are recommended to use the [RStudio IDE](https://www.rstudio.com/products/rstudio/), but you do not have to.

- Create a new R Markdown document from the menu `File -> New File -> R Markdown -> From Template -> Ninja Presentation`;<sup>1</sup>

--

- Click the `Knit` button to compile it;

--

- or use the [RStudio Addin](https://rstudio.github.io/rstudioaddins/)<sup>2</sup> "Infinite Moon Reader" to live preview the slides (every time you update and save the Rmd document, the slides will be automatically reloaded in RStudio Viewer.

.footnote[
[1] 中文用户请看[这份教程](https://slides.yihui.org/xaringan/zh-CN.html)

[2] See [#2](https://github.com/yihui/xaringan/issues/2) if you do not see the template or addin in RStudio.
]

---
background-image: url(`r xaringan:::karl`)
background-position: 50% 50%
class: center, bottom, inverse

# You only live once!

---

# Hello Ninja

As a presentation ninja, you certainly should not be satisfied by the "Hello World" example. You need to understand more about two things:

1. The [remark.js](https://remarkjs.com) library;

1. The **xaringan** package;

Basically **xaringan** injected the chakra of R Markdown (minus Pandoc) into **remark.js**. The slides are rendered by remark.js in the web browser, and the Markdown source needed by remark.js is generated from R Markdown (**knitr**).

---

# remark.js

You can see an introduction of remark.js from [its homepage](https://remarkjs.com). You should read the [remark.js Wiki](https://github.com/gnab/remark/wiki) at least once to know how to

- create a new slide (Markdown syntax<sup>*</sup> and slide properties);

- format a slide (e.g. text alignment);

- configure the slideshow;

- and use the presentation (keyboard shortcuts).

It is important to be familiar with remark.js before you can understand the options in **xaringan**.

.footnote[[*] It is different with Pandoc's Markdown! It is limited but should be enough for presentation purposes. Come on... You do not need a slide for the Table of Contents! Well, the Markdown support in remark.js [may be improved](https://github.com/gnab/remark/issues/142) in the future.]

---
background-image: url(`r xaringan:::karl`)
background-size: cover
class: center, bottom, inverse

# I was so happy to have discovered remark.js!

---
class: inverse, middle, center

# Using xaringan

---

# xaringan

Provides an R Markdown output format `xaringan::moon_reader` as a wrapper for remark.js, and you can use it in the YAML metadata, e.g.

```yaml
---
title: "A Cool Presentation"
output:
  xaringan::moon_reader:
    yolo: true
    nature:
      autoplay: 30000
---
```

See the help page `?xaringan::moon_reader` for all possible options that you can use.

---

# remark.js vs xaringan

Some differences between using remark.js (left) and using **xaringan** (right):

.pull-left[
1. Start with a boilerplate HTML file;

1. Plain Markdown;

1. Write JavaScript to autoplay slides;

1. Manually configure MathJax;

1. Highlight code with `*`;

1. Edit Markdown source and refresh browser to see updated slides;
]

.pull-right[
1. Start with an R Markdown document;

1. R Markdown (can embed R/other code chunks);

1. Provide an option `autoplay`;

1. MathJax just works;<sup>*</sup>

1. Highlight code with `{{}}`;

1. The RStudio addin "Infinite Moon Reader" automatically refreshes slides on changes;
]

.footnote[[*] Not really. See next page.]

---

# Math Expressions

You can write LaTeX math expressions inside a pair of dollar signs, e.g. &#36;\alpha+\beta$ renders $\alpha+\beta$. You can use the display style with double dollar signs:

```
$$\bar{X}=\frac{1}{n}\sum_{i=1}^nX_i$$
```

$$\bar{X}=\frac{1}{n}\sum_{i=1}^nX_i$$

Limitations:

1. The source code of a LaTeX math expression must be in one line, unless it is inside a pair of double dollar signs, in which case the starting `$$` must appear in the very beginning of a line, followed immediately by a non-space character, and the ending `$$` must be at the end of a line, led by a non-space character;

1. There should not be spaces after the opening `$` or before the closing `$`.

1. Math does not work on the title slide (see [#61](https://github.com/yihui/xaringan/issues/61) for a workaround).

---

# R Code

```{r comment='#'}
# a boring regression
fit = lm(dist ~ 1 + speed, data = cars)
coef(summary(fit))
dojutsu = c('地爆天星', '天照', '加具土命', '神威', '須佐能乎', '無限月読')
grep('天', dojutsu, value = TRUE)
```

---

# R Plots

```{r cars, fig.height=4, dev='svg'}
par(mar = c(4, 4, 1, .1))
plot(cars, pch = 19, col = 'darkgray', las = 1)
abline(fit, lwd = 2)
```

---

# Tables

If you want to generate a table, make sure it is in the HTML format (instead of Markdown or other formats), e.g.,

```{r}
knitr::kable(head(iris), format = 'html')
```

---

# HTML Widgets

I have not thoroughly tested HTML widgets against **xaringan**. Some may work well, and some may not. It is a little tricky.

Similarly, the Shiny mode (`runtime: shiny`) does not work. I might get these issues fixed in the future, but these are not of high priority to me. I never turn my presentation into a Shiny app. When I need to demonstrate more complicated examples, I just launch them separately. It is convenient to share slides with other people when they are plain HTML/JS applications.

See the next page for two HTML widgets.

---

```{r out.width='100%', fig.height=6, eval=require('leaflet')}
library(leaflet)
leaflet() %>% addTiles() %>% setView(-93.65, 42.0285, zoom = 17)
```

---

```{r eval=require('DT'), tidy=FALSE}
DT::datatable(
  head(iris, 10),
  fillContainer = FALSE, options = list(pageLength = 8)
)
```

---

# Some Tips

- Do not forget to try the `yolo` option of `xaringan::moon_reader`.

    ```yaml
    output:
      xaringan::moon_reader:
        yolo: true
    ```

---

# Some Tips

- Slides can be automatically played if you set the `autoplay` option under `nature`, e.g. go to the next slide every 30 seconds in a lightning talk:

    ```yaml
    output:
      xaringan::moon_reader:
        nature:
          autoplay: 30000
    ```

- If you want to restart the play after it reaches the last slide, you may set the sub-option `loop` to TRUE, e.g.,

    ```yaml
    output:
      xaringan::moon_reader:
        nature:
          autoplay:
            interval: 30000
            loop: true
    ```

---

# Some Tips

- A countdown timer can be added to every page of the slides using the `countdown` option under `nature`, e.g. if you want to spend one minute on every page when you give the talk, you can set:

    ```yaml
    output:
      xaringan::moon_reader:
        nature:
          countdown: 60000
    ```

    Then you will see a timer counting down from `01:00`, to `00:59`, `00:58`, ... When the time is out, the timer will continue but the time turns red.

---

# Some Tips

- The title slide is created automatically by **xaringan**, but it is just another remark.js slide added before your other slides.

    The title slide is set to `class: center, middle, inverse, title-slide` by default. You can change the classes applied to the title slide with the `titleSlideClass` option of `nature` (`title-slide` is always applied).

    ```yaml
    output:
      xaringan::moon_reader:
        nature:
          titleSlideClass: [top, left, inverse]
    ```

--

- If you'd like to create your own title slide, disable **xaringan**'s title slide with the `seal = FALSE` option of `moon_reader`.

    ```yaml
    output:
      xaringan::moon_reader:
        seal: false
    ```

---

# Some Tips

- There are several ways to build incremental slides. See [this presentation](https://slides.yihui.org/xaringan/incremental.html) for examples.

- The option `highlightLines: true` of `nature` will highlight code lines that start with `*`, or are wrapped in `{{ }}`, or have trailing comments `#<<`;

    ```yaml
    output:
      xaringan::moon_reader:
        nature:
          highlightLines: true
    ```

    See examples on the next page.

---

# Some Tips

.pull-left[
An example using a leading `*`:

    ```r
    if (TRUE) {
    ** message("Very important!")
    }
    ```
Output:
```r
if (TRUE) {
* message("Very important!")
}
```

This is invalid R code, so it is a plain fenced code block that is not executed.
]

.pull-right[
An example using `{{}}`:

````
`r ''````{r tidy=FALSE}
if (TRUE) {
*{{ message("Very important!") }}
}
```
````
Output:
```{r tidy=FALSE}
if (TRUE) {
{{ message("Very important!") }}
}
```

It is valid R code so you can run it. Note that `{{}}` can wrap an R expression of multiple lines.
]

---

# Some Tips

An example of using the trailing comment `#<<` to highlight lines:

````markdown
`r ''````{r tidy=FALSE}
library(ggplot2)
ggplot(mtcars) + 
  aes(mpg, disp) + 
  geom_point() +   #<<
  geom_smooth()    #<<
```
````

Output:

```{r tidy=FALSE, eval=FALSE}
library(ggplot2)
ggplot(mtcars) + 
  aes(mpg, disp) + 
  geom_point() +   #<<
  geom_smooth()    #<<
```

---

# Some Tips

When you enable line-highlighting, you can also use the chunk option `highlight.output` to highlight specific lines of the text output from a code chunk. For example, `highlight.output = TRUE` means highlighting all lines, and `highlight.output = c(1, 3)` means highlighting the first and third line.

````md
`r ''````{r, highlight.output=c(1, 3)}
head(iris)
```
````

```{r, highlight.output=c(1, 3), echo=FALSE}
head(iris)
```

Question: what does `highlight.output = c(TRUE, FALSE)` mean? (Hint: think about R's recycling of vectors)

---

# Some Tips

- To make slides work offline, you need to download a copy of remark.js in advance, because **xaringan** uses the online version by default (see the help page `?xaringan::moon_reader`).

- You can use `xaringan::summon_remark()` to download the latest or a specified version of remark.js. By default, it is downloaded to `libs/remark-latest.min.js`.

- Then change the `chakra` option in YAML to point to this file, e.g.

    ```yaml
    output:
      xaringan::moon_reader:
        chakra: libs/remark-latest.min.js
    ```

- If you used Google fonts in slides (the default theme uses _Yanone Kaffeesatz_, _Droid Serif_, and _Source Code Pro_), they won't work offline unless you download or install them locally. The Heroku app [google-webfonts-helper](https://google-webfonts-helper.herokuapp.com/fonts) can help you download fonts and generate the necessary CSS.

---

# Macros

- remark.js [allows users to define custom macros](https://github.com/yihui/xaringan/issues/80) (JS functions) that can be applied to Markdown text using the syntax `![:macroName arg1, arg2, ...]` or `![:macroName arg1, arg2, ...](this)`. For example, before remark.js initializes the slides, you can define a macro named `scale`:

    ```js
    remark.macros.scale = function (percentage) {
      var url = this;
      return '<img src="' + url + '" style="width: ' + percentage + '" />';
    };
    ```

    Then the Markdown text

    ```markdown
    ![:scale 50%](image.jpg)
    ```

    will be translated to

    ```html
    <img src="image.jpg" style="width: 50%" />
    ```

---

# Macros (continued)

- To insert macros in **xaringan** slides, you can use the option `beforeInit` under the option `nature`, e.g.,

    ```yaml
    output:
      xaringan::moon_reader:
        nature:
          beforeInit: "macros.js"
    ```

    You save your remark.js macros in the file `macros.js`.

- The `beforeInit` option can be used to insert arbitrary JS code before `remark.create()`. Inserting macros is just one of its possible applications.

---

# CSS

Among all options in `xaringan::moon_reader`, the most challenging but perhaps also the most rewarding one is `css`, because it allows you to customize the appearance of your slides using any CSS rules or hacks you know.

You can see the default CSS file [here](https://github.com/yihui/xaringan/blob/master/inst/rmarkdown/templates/xaringan/resources/default.css). You can completely replace it with your own CSS files, or define new rules to override the default. See the help page `?xaringan::moon_reader` for more information.

---

# CSS

For example, suppose you want to change the font for code from the default "Source Code Pro" to "Ubuntu Mono". You can create a CSS file named, say, `ubuntu-mono.css`:

```css
@import url(https://fonts.googleapis.com/css?family=Ubuntu+Mono:400,700,400italic);

.remark-code, .remark-inline-code { font-family: 'Ubuntu Mono'; }
```

Then set the `css` option in the YAML metadata:

```yaml
output:
  xaringan::moon_reader:
    css: ["default", "ubuntu-mono.css"]
```

Here I assume `ubuntu-mono.css` is under the same directory as your Rmd.

See [yihui/xaringan#83](https://github.com/yihui/xaringan/issues/83) for an example of using the [Fira Code](https://github.com/tonsky/FiraCode) font, which supports ligatures in program code.

---

# CSS (with Sass)

**xaringan** also supports Sass support via **rmarkdown**. Suppose you want to use the same color for different elements, e.g., first heading and bold text. You can create a `.scss` file, say `mytheme.scss`, using the [sass](https://sass-lang.com/) syntax with variables:

```scss
$mycolor: #ff0000; 
.remark-slide-content > h1 { color: $mycolor; }
.remark-slide-content strong { color: $mycolor; }
```

Then set the `css` option in the YAML metadata using this file placed under the same directory as your Rmd:

```yaml
output:
  xaringan::moon_reader:
    css: ["default", "mytheme.scss"]
```

This requires **rmarkdown** >= 2.8 and the [**sass**](https://rstudio.github.io/sass/) package. You can learn more about **rmarkdown** and **sass** support in [this blog post](https://blog.rstudio.com/2021/04/15/2021-spring-rmd-news/#sass-and-scss-support-for-html-based-output) and in [**sass** overview vignette](https://rstudio.github.io/sass/articles/sass.html).

---

# Themes

Don't want to learn CSS? Okay, you can use some user-contributed themes. A theme typically consists of two CSS files `foo.css` and `foo-fonts.css`, where `foo` is the theme name. Below are some existing themes:

```{r, R.options=list(width = 70)}
names(xaringan:::list_css())
```

---

# Themes

To use a theme, you can specify the `css` option as an array of CSS filenames (without the `.css` extensions), e.g.,

```yaml
output:
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts]
```

If you want to contribute a theme to **xaringan**, please read [this blog post](https://yihui.org/en/2017/10/xaringan-themes).

---
class: inverse, middle, center
background-image: url(https://upload.wikimedia.org/wikipedia/commons/3/39/Naruto_Shiki_Fujin.svg)
background-size: contain

# Naruto

---
background-image: url(https://upload.wikimedia.org/wikipedia/commons/b/be/Sharingan_triple.svg)
background-size: 100px
background-position: 90% 8%

# Sharingan

The R package name **xaringan** was derived<sup>1</sup> from **Sharingan**, a dōjutsu in the Japanese anime _Naruto_ with two abilities:

- the "Eye of Insight"

- the "Eye of Hypnotism"

I think a presentation is basically a way to communicate insights to the audience, and a great presentation may even "hypnotize" the audience.<sup>2,3</sup>

.footnote[
[1] In Chinese, the pronounciation of _X_ is _Sh_ /ʃ/ (as in _shrimp_). Now you should have a better idea of how to pronounce my last name _Xie_.

[2] By comparison, bad presentations only put the audience to sleep.

[3] Personally I find that setting background images for slides is a killer feature of remark.js. It is an effective way to bring visual impact into your presentations.
]

---

# Naruto terminology

The **xaringan** package borrowed a few terms from Naruto, such as

- [Sharingan](https://naruto.fandom.com/wiki/Sharingan) (写輪眼; the package name)

- The [moon reader](https://naruto.fandom.com/wiki/Moon_Reader) (月読; an attractive R Markdown output format)

- [Chakra](https://naruto.fandom.com/wiki/Chakra) (查克拉; the path to the remark.js library, which is the power to drive the presentation)

- [Nature transformation](https://naruto.fandom.com/wiki/Nature_Transformation) (性質変化; transform the chakra by setting different options)

- The [infinite moon reader](https://naruto.fandom.com/wiki/Infinite_Tsukuyomi) (無限月読; start a local web server to continuously serve your slides)

- The [summoning technique](https://naruto.fandom.com/wiki/Summoning_Technique) (download remark.js from the web)

You can click the links to know more about them if you want. The jutsu "Moon Reader" may seem a little evil, but that does not mean your slides are evil.

---

class: center

# Hand seals (印)

Press `h` or `?` to see the possible ninjutsu you can use in remark.js.

![](https://upload.wikimedia.org/wikipedia/commons/7/7e/Mudra-Naruto-KageBunshin.svg)

---

class: center, middle

# Thanks!

Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).

The chakra comes from [remark.js](https://remarkjs.com), [**knitr**](https://yihui.org/knitr/), and [R Markdown](https://rmarkdown.rstudio.com).

### 2021-09-06 draft.Rmd
추가 - 작업 단계 screenshot

@@
   - 프로젝트 관리
-  - R 코딩 습관
+  - 단위 테스트
   - 메타 프로그래밍
-  - 단위 테스트
+  - Refactoring
 
@@
 - 참고교재
+
 - 실무용 아님
+
 - 제한된 유연성
@@
 
+.pull-left[
+
 - RStudio IDE
+
 - `{devtools}`, `{usethis}`, `{testthat}`, etc.
+
 - Git and GitHub
 
-```{r r-packages-book, echo = FALSE, fig.cap="R Packages / O'Reilly / Hadley Wickham", out.width='25%', fig.align='center'}
+]
+
+--
+
+.pull-right[
+
+```{r r-packages-book, echo = FALSE, fig.cap="R Packages / O'Reilly / Hadley Wickham", out.width='40%', fig.align='center'}
 knitr::include_graphics("images/r-pkgs-cover.png")
+```
+
+]
+
+---
+
+# R 스크립트 분류 및 재구성
+
+- R 스크립트를 목적별로 분류하여 각 목적에 맞는 폴더에 저장
+
+.pull-left[
+
+```{r r-script-before, echo = FALSE, fig.cap="작업 이전 R Markdown 파일", out.width='80%', fig.align='center'}
+knitr::include_graphics("images/r-script-before.png")
+```
+
+]
+
+--
+
+.pull-right[
+
+```{r r-script-after-data-raw, echo = FALSE, out.width='70%', fig.cap="데이터 생성: data-raw/ 폴더", fig.align='center'}
+knitr::include_graphics("images/r-after-data-raw.png")
+```
+
+```{r r-script-after-function-r, echo = FALSE, out.width='70%', fig.cap="함수 정의: R/ 폴더", fig.align='center'}
+knitr::include_graphics("images/r-after-function-r.png")
+```
+
+```{r r-script-after-vignettes, echo = FALSE, out.width='70%', fig.cap="데이터 분석 문서: vignettes/ 폴더", fig.align='center'}
+knitr::include_graphics("images/r-after-vignettes.png")
+```
+
+]
+
+---
+
+# 함수 도움말 문서 작성
+
+- `{roxygen2}`를 이용한 함수 도움말 작성
+
+```{r r-function-manual, echo = FALSE, fig.cap="함수 도움말 작성 및 컴파일된 도움말 파일", out.width='49%', fig.align='center', fig.show='hold'}
+knitr::include_graphics(c(
+  "images/r-function-manual.png",
+  "images/r-function-manual-rendered.png"
+))
 ```
@@
 
-# 문서 내 R 스크립트 분류
+# 단위 테스트 작성
 
-- 데이터 생성 --> data-raw/
-- 함수 생성 --> R/
-- 데이터 분석  --> vignettes/
+- `{testthat}`을 이용한 예제 결과 재현성 테스트
+
+.pull-left[
+
+```{r unit-test-book-example, echo = FALSE, fig.cap="단위 테스트 사용할 예제", out.width='50%', fig.align='center'}
+knitr::include_graphics("images/unit-test-book-example.png")
+```
+
+]
+
+--
+
+.pull-right[
+
+```{r unit-test-script, echo = FALSE, out.width='70%', fig.cap="단위 테스트 작성: tests/ 폴더", fig.align='center'}
+knitr::include_graphics("images/unit-test-script.png")
+```
+
+```{r unit-test-result, echo = FALSE, out.width='50%', fig.cap="단위 테스트 결과", fig.align='center'}
+knitr::include_graphics("images/unit-test-result.png")
+```
+
+]
+
+---
+
+# 패키지 사이트 생성
+
+- `{pkgdown}`을 이용한 패키지 웹사이트 생성
+
+```{r pkgdown-site, echo = FALSE, out.width='50%', fig.cap="패키지 웹사이트", fig.align='center'}
+knitr::include_graphics("images/pkgdown-site.gif")
+```
+
+---
+class: inverse, center, middle
+
+# 데이터 분석가에게 유용한 개발자 관점
+
+---
+
+# 프로젝트 관리
+
+- 일관된 프로젝트 폴더 구조
+
+- Package dependency
+
+- Git
+
+---
+
+# 단위 테스트
+
+-
+
+---
+
+# 리팩토링 (Refactoring)
+
+- 함수화
+
+- 메타 프로그래밍
+
+-
 

### 2021-09-06 draft.Rmd
슬라이드 제목 한글화
수정 - outline 깔끔하게 공백 추가

@@
 1. 데이터 분석 문서화에서 패키지 개발로 전환
+
   - 배경
+
   - 작업 과정
+
 2. 데이터 분석가에게 유용한 개발자 관점
+
   - 프로젝트 관리
+
   - 단위 테스트
+
   - 메타 프로그래밍
-  - Refactoring
+
+  - 리펙토링
 
@@
 
-# 현실적인 scope 정의
+# 현실적인 작업범위 및 기대수준 정의
 
@@
 ---
-
-# Hello World
-
-Install the **xaringan** package from [Github](https://github.com/yihui/xaringan):
-
-```{r eval=FALSE, tidy=FALSE}
-remotes::install_github("yihui/xaringan")
-```
-
---
-
-You are recommended to use the [RStudio IDE](https://www.rstudio.com/products/rstudio/), but you do not have to.
-
-- Create a new R Markdown document from the menu `File -> New File -> R Markdown -> From Template -> Ninja Presentation`;<sup>1</sup>
-
---
-
-- Click the `Knit` button to compile it;
-
---
-
-- or use the [RStudio Addin](https://rstudio.github.io/rstudioaddins/)<sup>2</sup> "Infinite Moon Reader" to live preview the slides (every time you update and save the Rmd document, the slides will be automatically reloaded in RStudio Viewer.
-
-.footnote[
-[1] 中文用户请看[这份教程](https://slides.yihui.org/xaringan/zh-CN.html)
-
-[2] See [#2](https://github.com/yihui/xaringan/issues/2) if you do not see the template or addin in RStudio.
-]
-
----
-background-image: url(`r xaringan:::karl`)
-background-position: 50% 50%
-class: center, bottom, inverse
-
-# You only live once!
-
----
-
-# Hello Ninja
-
-As a presentation ninja, you certainly should not be satisfied by the "Hello World" example. You need to understand more about two things:
-
-1. The [remark.js](https://remarkjs.com) library;
-
-1. The **xaringan** package;
-
-Basically **xaringan** injected the chakra of R Markdown (minus Pandoc) into **remark.js**.
-The slides are rendered by remark.js in the web browser, and the Markdown source needed by remark.js is generated from R Markdown (**knitr**).
-
----
-
-# remark.js
-
-You can see an introduction of remark.js from [its homepage](https://remarkjs.com). You should read the [remark.js Wiki](https://github.com/gnab/remark/wiki) at least once to know how to
-
-- create a new slide (Markdown syntax<sup>*</sup> and slide properties);
-
-- format a slide (e.g. text alignment);
-
-- configure the slideshow;
-
-- and use the presentation (keyboard shortcuts).
-
-It is important to be familiar with remark.js before you can understand the options in **xaringan**.
-
-.footnote[[*] It is different with Pandoc's Markdown!
-It is limited but should be enough for presentation purposes.
-Come on...
-You do not need a slide for the Table of Contents!
-Well, the Markdown support in remark.js [may be improved](https://github.com/gnab/remark/issues/142) in the future.]
-
----
-background-image: url(`r xaringan:::karl`)
-background-size: cover
-class: center, bottom, inverse
-
-# I was so happy to have discovered remark.js!
-
----
-class: inverse, middle, center
-
-# Using xaringan
-
----
-
-# xaringan
-
-Provides an R Markdown output format `xaringan::moon_reader` as a wrapper for remark.js, and you can use it in the YAML metadata, e.g.
-
-```yaml
----
-title: "A Cool Presentation"
-output:
-  xaringan::moon_reader:
-    yolo: true
-    nature:
-      autoplay: 30000
----
-```
-
-See the help page `?xaringan::moon_reader` for all possible options that you can use.
-
----
-
-# remark.js vs xaringan
-
-Some differences between using remark.js (left) and using **xaringan** (right):
-
-.pull-left[
-1. Start with a boilerplate HTML file;
-
-1. Plain Markdown;
-
-1. Write JavaScript to autoplay slides;
-
-1. Manually configure MathJax;
-
-1. Highlight code with `*`;
-
-1. Edit Markdown source and refresh browser to see updated slides;
-]
-
-.pull-right[
-1. Start with an R Markdown document;
-
-1. R Markdown (can embed R/other code chunks);
-
-1. Provide an option `autoplay`;
-
-1. MathJax just works;<sup>*</sup>
-
-1. Highlight code with `{{}}`;
-
-1. The RStudio addin "Infinite Moon Reader" automatically refreshes slides on changes;
-]
-
-.footnote[[*] Not really. See next page.]
-
----
-
-# Math Expressions
-
-You can write LaTeX math expressions inside a pair of dollar signs, e.g. &#36;\alpha+\beta$ renders $\alpha+\beta$. You can use the display style with double dollar signs:
-
-```
-$$\bar{X}=\frac{1}{n}\sum_{i=1}^nX_i$$
-```
-
-$$\bar{X}=\frac{1}{n}\sum_{i=1}^nX_i$$
-
-Limitations:
-
-1.
-The source code of a LaTeX math expression must be in one line, unless it is inside a pair of double dollar signs, in which case the starting `$$` must appear in the very beginning of a line, followed immediately by a non-space character, and the ending `$$` must be at the end of a line, led by a non-space character;
-
-1. There should not be spaces after the opening `$` or before the closing `$`.
-
-1. Math does not work on the title slide (see [#61](https://github.com/yihui/xaringan/issues/61) for a workaround).
-
----
-
-# R Code
-
-```{r comment='#'}
-# a boring regression
-fit = lm(dist ~ 1 + speed, data = cars)
-coef(summary(fit))
-dojutsu = c('地爆天星', '天照', '加具土命', '神威', '須佐能乎', '無限月読')
-grep('天', dojutsu, value = TRUE)
-```
-
----
-
-# R Plots
-
-```{r cars, fig.height=4, dev='svg'}
-par(mar = c(4, 4, 1, .1))
-plot(cars, pch = 19, col = 'darkgray', las = 1)
-abline(fit, lwd = 2)
-```
-
----
-
-# Tables
-
-If you want to generate a table, make sure it is in the HTML format (instead of Markdown or other formats), e.g.,
-
-```{r}
-knitr::kable(head(iris), format = 'html')
-```
-
----
-
-# HTML Widgets
-
-I have not thoroughly tested HTML widgets against **xaringan**. Some may work well, and some may not. It is a little tricky.
-
-Similarly, the Shiny mode (`runtime: shiny`) does not work.
-I might get these issues fixed in the future, but these are not of high priority to me.
-I never turn my presentation into a Shiny app.
-When I need to demonstrate more complicated examples, I just launch them separately.
-It is convenient to share slides with other people when they are plain HTML/JS applications.
-
-See the next page for two HTML widgets.
-
----
-
-```{r out.width='100%', fig.height=6, eval=require('leaflet')}
-library(leaflet)
-leaflet() %>% addTiles() %>% setView(-93.65, 42.0285, zoom = 17)
-```
-
----
-
-```{r eval=require('DT'), tidy=FALSE}
-DT::datatable(
-  head(iris, 10),
-  fillContainer = FALSE, options = list(pageLength = 8)
-)
-```
-
----
-
-# Some Tips
-
-- Do not forget to try the `yolo` option of `xaringan::moon_reader`.
-
-    ```yaml
-    output:
-      xaringan::moon_reader:
-        yolo: true
-    ```
-
----
-
-# Some Tips
-
-- Slides can be automatically played if you set the `autoplay` option under `nature`, e.g. go to the next slide every 30 seconds in a lightning talk:
-
-    ```yaml
-    output:
-      xaringan::moon_reader:
-        nature:
-          autoplay: 30000
-    ```
-
-- If you want to restart the play after it reaches the last slide, you may set the sub-option `loop` to TRUE, e.g.,
-
-    ```yaml
-    output:
-      xaringan::moon_reader:
-        nature:
-          autoplay:
-            interval: 30000
-            loop: true
-    ```
-
----
-
-# Some Tips
-
-- A countdown timer can be added to every page of the slides using the `countdown` option under `nature`, e.g. if you want to spend one minute on every page when you give the talk, you can set:
-
-    ```yaml
-    output:
-      xaringan::moon_reader:
-        nature:
-          countdown: 60000
-    ```
-
-    Then you will see a timer counting down from `01:00`, to `00:59`, `00:58`, ... When the time is out, the timer will continue but the time turns red.
-
----
-
-# Some Tips
-
-- The title slide is created automatically by **xaringan**, but it is just another remark.js slide added before your other slides.
-
-    The title slide is set to `class: center, middle, inverse, title-slide` by default.
-You can change the classes applied to the title slide with the `titleSlideClass` option of `nature` (`title-slide` is always applied).
-
-    ```yaml
-    output:
-      xaringan::moon_reader:
-        nature:
-          titleSlideClass: [top, left, inverse]
-    ```
-
---
-
-- If you'd like to create your own title slide, disable **xaringan**'s title slide with the `seal = FALSE` option of `moon_reader`.
-
-    ```yaml
-    output:
-      xaringan::moon_reader:
-        seal: false
-    ```
-
----
-
-# Some Tips
-
-- There are several ways to build incremental slides. See [this presentation](https://slides.yihui.org/xaringan/incremental.html) for examples.
-
-- The option `highlightLines: true` of `nature` will highlight code lines that start with `*`, or are wrapped in `{{ }}`, or have trailing comments `#<<`;
-
-    ```yaml
-    output:
-      xaringan::moon_reader:
-        nature:
-          highlightLines: true
-    ```
-
-    See examples on the next page.
-
----
-
-# Some Tips
-
-.pull-left[
-An example using a leading `*`:
-
-    ```r
-    if (TRUE) {
-    ** message("Very important!")
-    }
-    ```
-Output:
-```r
-if (TRUE) {
-* message("Very important!")
-}
-```
-
-This is invalid R code, so it is a plain fenced code block that is not executed.
-]
-
-.pull-right[
-An example using `{{}}`:
-
-````
-`r ''````{r tidy=FALSE}
-if (TRUE) {
-*{{ message("Very important!") }}
-}
-```
-````
-Output:
-```{r tidy=FALSE}
-if (TRUE) {
-{{ message("Very important!") }}
-}
-```
-
-It is valid R code so you can run it. Note that `{{}}` can wrap an R expression of multiple lines.
-]
-
----
-
-# Some Tips
-
-An example of using the trailing comment `#<<` to highlight lines:
-
-````markdown
-`r ''````{r tidy=FALSE}
-library(ggplot2)
-ggplot(mtcars) + 
-  aes(mpg, disp) + 
-  geom_point() +   #<<
-  geom_smooth()    #<<
-```
-````
-
-Output:
-
-```{r tidy=FALSE, eval=FALSE}
-library(ggplot2)
-ggplot(mtcars) + 
-  aes(mpg, disp) + 
-  geom_point() +   #<<
-  geom_smooth()    #<<
-```
-
----
-
-# Some Tips
-
-When you enable line-highlighting, you can also use the chunk option `highlight.output` to highlight specific lines of the text output from a code chunk.
-For example, `highlight.output = TRUE` means highlighting all lines, and `highlight.output = c(1, 3)` means highlighting the first and third line.
-
-````md
-`r ''````{r, highlight.output=c(1, 3)}
-head(iris)
-```
-````
-
-```{r, highlight.output=c(1, 3), echo=FALSE}
-head(iris)
-```
-
-Question: what does `highlight.output = c(TRUE, FALSE)` mean? (Hint: think about R's recycling of vectors)
-
----
-
-# Some Tips
-
-- To make slides work offline, you need to download a copy of remark.js in advance, because **xaringan** uses the online version by default (see the help page `?xaringan::moon_reader`).
-
-- You can use `xaringan::summon_remark()` to download the latest or a specified version of remark.js. By default, it is downloaded to `libs/remark-latest.min.js`.
-
-- Then change the `chakra` option in YAML to point to this file, e.g.
-
-    ```yaml
-    output:
-      xaringan::moon_reader:
-        chakra: libs/remark-latest.min.js
-    ```
-
-- If you used Google fonts in slides (the default theme uses _Yanone Kaffeesatz_, _Droid Serif_, and _Source Code Pro_), they won't work offline unless you download or install them locally.
-The Heroku app [google-webfonts-helper](https://google-webfonts-helper.herokuapp.com/fonts) can help you download fonts and generate the necessary CSS.
-
----
-
-# Macros
-
-- remark.js [allows users to define custom macros](https://github.com/yihui/xaringan/issues/80) (JS functions) that can be applied to Markdown text using the syntax `![:macroName arg1, arg2, ...]` or `![:macroName arg1, arg2, ...](this)`.
-For example, before remark.js initializes the slides, you can define a macro named `scale`:
-
-    ```js
-    remark.macros.scale = function (percentage) {
-      var url = this;
-      return '<img src="' + url + '" style="width: ' + percentage + '" />';
-    };
-    ```
-
-    Then the Markdown text
-
-    ```markdown
-    ![:scale 50%](image.jpg)
-    ```
-
-    will be translated to
-
-    ```html
-    <img src="image.jpg" style="width: 50%" />
-    ```
-
----
-
-# Macros (continued)
-
-- To insert macros in **xaringan** slides, you can use the option `beforeInit` under the option `nature`, e.g.,
-
-    ```yaml
-    output:
-      xaringan::moon_reader:
-        nature:
-          beforeInit: "macros.js"
-    ```
-
-    You save your remark.js macros in the file `macros.js`.
-
-- The `beforeInit` option can be used to insert arbitrary JS code before `remark.create()`. Inserting macros is just one of its possible applications.
-
----
-
-# CSS
-
-Among all options in `xaringan::moon_reader`, the most challenging but perhaps also the most rewarding one is `css`, because it allows you to customize the appearance of your slides using any CSS rules or hacks you know.
-
-You can see the default CSS file [here](https://github.com/yihui/xaringan/blob/master/inst/rmarkdown/templates/xaringan/resources/default.css).
-You can completely replace it with your own CSS files, or define new rules to override the default.
-See the help page `?xaringan::moon_reader` for more information.
-
----
-
-# CSS
-
-For example, suppose you want to change the font for code from the default "Source Code Pro" to "Ubuntu Mono". You can create a CSS file named, say, `ubuntu-mono.css`:
-
-```css
-@import url(https://fonts.googleapis.com/css?family=Ubuntu+Mono:400,700,400italic);
-
-.remark-code, .remark-inline-code { font-family: 'Ubuntu Mono'; }
-```
-
-Then set the `css` option in the YAML metadata:
-
-```yaml
-output:
-  xaringan::moon_reader:
-    css: ["default", "ubuntu-mono.css"]
-```
-
-Here I assume `ubuntu-mono.css` is under the same directory as your Rmd.
-
-See [yihui/xaringan#83](https://github.com/yihui/xaringan/issues/83) for an example of using the [Fira Code](https://github.com/tonsky/FiraCode) font, which supports ligatures in program code.
-
----
-
-# CSS (with Sass)
-
-**xaringan** also supports Sass support via **rmarkdown**.
-Suppose you want to use the same color for different elements, e.g., first heading and bold text.
-You can create a `.scss` file, say `mytheme.scss`, using the [sass](https://sass-lang.com/) syntax with variables:
-
-```scss
-$mycolor: #ff0000; 
-.remark-slide-content > h1 { color: $mycolor; }
-.remark-slide-content strong { color: $mycolor; }
-```
-
-Then set the `css` option in the YAML metadata using this file placed under the same directory as your Rmd:
-
-```yaml
-output:
-  xaringan::moon_reader:
-    css: ["default", "mytheme.scss"]
-```
-
-This requires **rmarkdown** >= 2.8 and the [**sass**](https://rstudio.github.io/sass/) package.
-You can learn more about **rmarkdown** and **sass** support in [this blog post](https://blog.rstudio.com/2021/04/15/2021-spring-rmd-news/#sass-and-scss-support-for-html-based-output) and in [**sass** overview vignette](https://rstudio.github.io/sass/articles/sass.html).
-
----
-
-# Themes
-
-Don't want to learn CSS?
-Okay, you can use some user-contributed themes.
-A theme typically consists of two CSS files `foo.css` and `foo-fonts.css`, where `foo` is the theme name.
-Below are some existing themes:
-
-```{r, R.options=list(width = 70)}
-names(xaringan:::list_css())
-```
-
----
-
-# Themes
-
-To use a theme, you can specify the `css` option as an array of CSS filenames (without the `.css` extensions), e.g.,
-
-```yaml
-output:
-  xaringan::moon_reader:
-    css: [default, metropolis, metropolis-fonts]
-```
-
-If you want to contribute a theme to **xaringan**, please read [this blog post](https://yihui.org/en/2017/10/xaringan-themes).
-
----
-class: inverse, middle, center
-background-image: url(https://upload.wikimedia.org/wikipedia/commons/3/39/Naruto_Shiki_Fujin.svg)
-background-size: contain
-
-# Naruto
-
----
-background-image: url(https://upload.wikimedia.org/wikipedia/commons/b/be/Sharingan_triple.svg)
-background-size: 100px
-background-position: 90% 8%
-
-# Sharingan
-
-The R package name **xaringan** was derived<sup>1</sup> from **Sharingan**, a dōjutsu in the Japanese anime _Naruto_ with two abilities:
-
-- the "Eye of Insight"
-
-- the "Eye of Hypnotism"
-
-I think a presentation is basically a way to communicate insights to the audience, and a great presentation may even "hypnotize" the audience.<sup>2,3</sup>
-
-.footnote[
-[1] In Chinese, the pronounciation of _X_ is _Sh_ /ʃ/ (as in _shrimp_). Now you should have a better idea of how to pronounce my last name _Xie_.
-
-[2] By comparison, bad presentations only put the audience to sleep.
-
-[3] Personally I find that setting background images for slides is a killer feature of remark.js. It is an effective way to bring visual impact into your presentations.
-]
-
----
-
-# Naruto terminology
-
-The **xaringan** package borrowed a few terms from Naruto, such as
-
-- [Sharingan](https://naruto.fandom.com/wiki/Sharingan) (写輪眼; the package name)
-
-- The [moon reader](https://naruto.fandom.com/wiki/Moon_Reader) (月読; an attractive R Markdown output format)
-
-- [Chakra](https://naruto.fandom.com/wiki/Chakra) (查克拉; the path to the remark.js library, which is the power to drive the presentation)
-
-- [Nature transformation](https://naruto.fandom.com/wiki/Nature_Transformation) (性質変化; transform the chakra by setting different options)
-
-- The [infinite moon reader](https://naruto.fandom.com/wiki/Infinite_Tsukuyomi) (無限月読; start a local web server to continuously serve your slides)
-
-- The [summoning technique](https://naruto.fandom.com/wiki/Summoning_Technique) (download remark.js from the web)
-
-You can click the links to know more about them if you want. The jutsu "Moon Reader" may seem a little evil, but that does not mean your slides are evil.
-
----
-
-class: center
-
-# Hand seals (印)
-
-Press `h` or `?` to see the possible ninjutsu you can use in remark.js.
-
-![](https://upload.wikimedia.org/wikipedia/commons/7/7e/Mudra-Naruto-KageBunshin.svg)
-
----
-
-class: center, middle
-
-# Thanks!
-
-Slides created via the R package [**xaringan**](https://github.com/yihui/xaringan).
-
-The chakra comes from [remark.js](https://remarkjs.com), [**knitr**](https://yihui.org/knitr/), and [R Markdown](https://rmarkdown.rstudio.com).

### 2021-09-12 draft.Rmd
apply conference template

@@
 author: "이영록"
-date: "2021/11/19 (updated: `r Sys.Date()`)"
+date: "2021-11-19"
 output:
   xaringan::moon_reader:
+    css: 
+      - [default]
+      - css/rconf.css
+      - css/rconf-fonts.css
     lib_dir: libs
+    seal: false
     nature:
-      highlightStyle: github
+      highlightStyle: googlecode
       highlightLines: true
+      highlightLanguage: ["r"]
       countIncrementalSlides: false
-      ratio: '16:9'      
+      ratio: "16:9"
+    includes:
+      after_body: [css/rconf-logo.html]
+editor_options: 
+  chunk_output_type: console
 ---
 
-```{r setup, include=FALSE}
-options(htmltools.dir.version = FALSE)
+```{r setup, include = FALSE}
+library(gt)
+library(ggplot2)
+library(dplyr)
+
+options(
+  htmltools.dir.version = FALSE, 
+  htmltools.preserve.raw = FALSE,
+  tibble.width = 60, tibble.print_min = 6,
+  crayon.enabled = TRUE
+)
+
+knitr::opts_chunk$set(
+  echo = FALSE,
+  message = FALSE,
+  warning = FALSE,
+  comment = "",
+  digits = 3,
+  tidy = FALSE,
+  prompt = FALSE,
+  fig.align = 'center',
+  # fig.width = 7.252,
+  # fig.height = 4,
+  dpi = 600
+)
+
+# uncomment the following lines if you want to use the NHS-R theme colours by default
+# scale_fill_continuous <- partial(scale_fill_nhs, discrete = FALSE)
+# scale_fill_discrete <- partial(scale_fill_nhs, discrete = TRUE)
+# scale_colour_continuous <- partial(scale_colour_nhs, discrete = FALSE)
+# scale_colour_discrete <- partial(scale_colour_nhs, discrete = TRUE)
 ```
 
+class: title-slide, left, bottom
+
+# `r rmarkdown::metadata$title`
+----
+### `r rmarkdown::metadata$author`
+### `r rmarkdown::metadata$date`
+
+---
 class: inverse, left, middle
+
+# 발표 개요
+
+----
 
@@
 
-  - 단위 테스트
+  - 변화 관리
 
-  - 메타 프로그래밍
-
-  - 리펙토링
+  - 코드의 재사용성 향상
 
@@
 ```
+
+---
+
+# 분석 예제: 선형 분류분석
+
+.pull-left[
+
+```{r lda-example, echo=FALSE, out.width='49%', fig.align='center', fig.show='hold'}
+data(binaryclass2, package = "dmtr")
+
+binaryclass2 %>% 
+  gt::gt()
+```
+
+]
+
+.pull-right[
+
+```{r lda-example-plot, echo=FALSE, out.width='90%', fig.align='center'}
+binaryclass2 %>% 
+  ggplot(aes(x = x1, y = x2, color = class)) +
+  geom_abline(slope = 0.9781, intercept = 0.3414, 
+              color = "grey20", linetype = "dashed") +
+  geom_point(size = 5) +
+  geom_text(aes(label = class), vjust = -1, size = 5) +
+  scale_x_continuous(
+    breaks = seq(1, 10, by = 1),
+    minor_breaks = NULL,
+    expand = expansion(mult = 0.05)
+  ) +
+  scale_y_continuous(
+    breaks = seq(1, 10, by = 1),
+    minor_breaks = NULL,
+    expand = expansion(mult = 0.05)
+  ) +
+  scale_color_manual(
+    values = c(`1` = "steelblue", `2` = "firebrick")
+  ) +
+  theme_light() +
+  theme(
+    legend.position = "none",
+    text = element_text(size = 20),
+    aspect.ratio = 1
+  )
+```
+
+]
 
@@
 
-- 알고리즘의 단계별 결과를 교재 내용대로 보여주기 위해 함수를 세분화할 필요
+- 기존 R 패키지가 교재에 설명된 알고리즘과 정확히 일치하지 않는 부분에 대한 혼동을 줄일 필요
+
+.pull-left[
+
+```{r lda-book-example, echo = FALSE, fig.cap="교재 예제", out.width='70%', fig.align='center'}
+knitr::include_graphics("images/lda-book-example-plot.png")
+```
+
+]
+
+.pull-right[
+
+```{r lda-r-example, echo = FALSE, out.width='50%', fig.cap="{MASS} 패키지 결과", fig.align='center'}
+knitr::include_graphics("images/lda-r-example.png")
+```
+
+]
 
@@
 
-- 기존 R 패키지가 교재에 설명된 알고리즘과 정확히 일치하지 않는 부분에 대한 혼동을 줄일 필요
+- 알고리즘의 단계별 결과를 교재 내용대로 보여주기 위해 함수를 세분화할 필요
 
@@
 
-- Package dependency
+  + **data-raw/**: 데이터 생성/쿼리 R 스크립트 (.R)
 
-- Git
+  + **data/**: 생성된 데이터 파일 (.rda)
+
+  + **R/**: 재사용을 위해 모듈화된 데이터 분석 함수 (.R)
+
+  + **vignettes/**: 데이터 분석 문서 (.Rmd)
 
@@
 
-# 단위 테스트
+# 변화 관리
 
--
+- 단위 테스트
+  + 함수 구현의 수정, 패키지 업데이트 등의 변경사항 발생 후, 데이터 분석 결과가  옳게 나타나는지 확인하고, 오류 발생 시 빠르게 오류를 찾아 수정
+  + 단위 테스트 생성 후, 데이터 분석 함수의 수정을 보다 적극적으로 자주 수행할 수 있어 리팩토링(refactoring) 생산성 증가
+
+- 네임스페이스 명시화
+  + 함수 콜 수행 시 네임스페이스 충돌 방지
+      + `library(MASS); library(dplyr); tibble(x = 1:2, y = 2:3) %>% select(x)`
+      + `library(dplyr); library(MASS); tibble(x = 1:2, y = 2:3) %>% select(x)`
+      +  `dplyr::tibble(x = 1:2, y = 2:3) %>% dplyr::select(x)`
 
@@
 
-# 리팩토링 (Refactoring)
-
-- 함수화
+# 코드의 재사용성 향상
 

### 2021-09-13 draft.Rmd
수정 - 코드 screen shot을 실제 code chunk로 대체하여 가독성 높임

@@
 library(dplyr)
+library(dmtr)
 
@@
 
-# 데이터마이닝 기법과 응용
+# `{bookdown}`을 이용한 R 예제 책 만들기
+
+.pull-left[
 
@@
 ```
+
+]
+
+.pull-right[
+
+```{r online-book-webshot, include=FALSE, cache=TRUE}
+webshot::webshot(
+  url = "https://youngroklee-ml.github.io/data-mining-book/index.html",
+  file = "images/onlinebook-main.jpg",
+  zoom = 3
+)
+```
+
+```{r online-book, echo = FALSE, fig.cap="데이터마이닝 with R / 전치혁, 이혜선, 이종석, 이영록 / http://youngroklee-ml.github.io/data-mining-book/", out.height='400px', fig.align='center'}
+knitr::include_graphics("images/onlinebook-main.jpg")
+```
+
+]
 
@@
 ---
+layout: true
 
-# `{bookdown}`을 이용한 R 예제 책 만들기
+# 교재용 R 패키지 작성의 필요성
 
-```{r online-book-webshot, include=FALSE, cache=TRUE}
-webshot::webshot(
-  url = "https://youngroklee-ml.github.io/data-mining-book/index.html",
-  file = "images/onlinebook-main.jpg",
-  zoom = 3
-)
+- 알고리즘의 단계별 결과를 교재 내용대로 보여주기 위해 함수를 세분화할 필요
+
+.pull-left[
+
+```{r book-covarince, echo = FALSE, fig.cap="교재 예제", out.width='80%', fig.align='center'}
+knitr::include_graphics("images/book-covariance.png")
 ```
 
-```{r online-book, echo = FALSE, fig.cap="데이터마이닝 with R / Published with bookdown / 전치혁, 이혜선, 이종석, 이영록 공저", out.height='400px', fig.align='center'}
-knitr::include_graphics("images/onlinebook-main.jpg")
+]
+
+---
+
+.pull-right[
+
+```{r lda-mass-summary, echo = TRUE, out.width='50%', fig.cap="{MASS} 패키지 결과", fig.align='center'}
+library(MASS)
+MASS::lda(class ~ x1 + x2, data = binaryclass2)
+```
+
+]
+
+---
+
+.pull-right[
+
+```{r lda-dmtr-group-variance, echo = TRUE}
+library(dmtr)
+group_variance(binaryclass2, class, x1:x2)
+```
+
+]
+
+---
+
+.pull-right[
+
+```{r lda-dmtr-pooled-variance, echo = TRUE}
+library(dmtr)
+pooled_variance(binaryclass2, class, x1:x2)
+```
+
+]
+
+---
+layout: true
+
+# 교재용 R 패키지 작성의 필요성
+
+- 기존 R 패키지 수행 결과가 교재의 예제와 정확히 일치하지 않는 부분에 대한 혼동을 줄일 필요
+
+```{r lda-book-example, echo = FALSE, fig.cap="교재 예제", out.width='50%', fig.align='center'}
+knitr::include_graphics("images/lda-book-example-function.png")
 ```
@@
 
-# 교재용 R 패키지 작성의 필요성
+--
+
+```{r lda-mass-result, echo = TRUE, out.width='50%', fig.cap="{MASS} 패키지 결과", fig.align='center'}
+fit <- MASS::lda(class ~ x1 + x2, data = binaryclass2)
+coef(fit) %>% drop()
+```
 
@@
 
-- 기존 R 패키지가 교재에 설명된 알고리즘과 정확히 일치하지 않는 부분에 대한 혼동을 줄일 필요
-
-.pull-left[
-
-```{r lda-book-example, echo = FALSE, fig.cap="교재 예제", out.width='70%', fig.align='center'}
-knitr::include_graphics("images/lda-book-example-plot.png")
+```{r lda-mass-result-transform, echo = TRUE, out.width='50%', fig.cap="{MASS} 패키지 결과 변환", fig.align='center'}
+scaling_adjust <- 2.201359; sign_adjust <- -1;
+(coef(fit) * scaling_adjust * sign_adjust) %>% drop()
 ```
 
-]
+---
 
-.pull-right[
-
-```{r lda-r-example, echo = FALSE, out.width='50%', fig.cap="{MASS} 패키지 결과", fig.align='center'}
-knitr::include_graphics("images/lda-r-example.png")
+```{r lda-dmtr-result, echo = TRUE}
+dmtr::fisher_ld(binaryclass2, class, x1:x2)
 ```
 
-]
+---
+layout: false
 
---
+# 교재용 R 패키지 작성의 필요성
 
-- 알고리즘의 단계별 결과를 교재 내용대로 보여주기 위해 함수를 세분화할 필요
+- 분석 알고리즘을 교재와 일치하도록 함수로 구현
 
---
+```r
+fisher_ld <- function(.data, .group_var, .xvar) {
+  .group_var <- rlang::enquo(.group_var)
+  .xvar <- rlang::enquo(.xvar)
 
-- 사용된 기존 R 패키지의 버전업에 따른 교재의 R 스크립트 수정을 줄일 필요
+* mu_hat <- group_mean(.data, !!.group_var, !!.xvar)
+
+  if (length(mu_hat) != 2L) {
+    stop("number of levels of .group_var must be 2.")
+  }
+
+* sigma_hat <- pooled_variance(.data, !!.group_var, !!.xvar)
+
+* res <- solve(sigma_hat) %*% purrr::reduce(mu_hat, `-`) %>% drop()
+
+  attr(res, "group") <- attr(mu_hat, "group")
+
+  return(res)
+}
+```
+
+```{r, include=FALSE}
+detach("package:MASS", character.only = TRUE)
+```
 
@@
 class: inverse, center, middle
+layout: false
 
@@
 ---
+layout: true
 
@@
 
+---
+
 - R 스크립트를 목적별로 분류하여 각 목적에 맞는 폴더에 저장
+  + 데이터 생성: data-raw/ 폴더에 .R 파일로 저장
+  + 함수 정의: R/ 폴더에 .R 파일로 저장
+  + 데이터 분석: vignettes/ 폴더에 .Rmd 파일로 저장
 
-.pull-left[
+```r
+binaryclass2 <- tibble::tibble(
+  id = c(1:9),
+  x1 = c(5, 4, 7, 8, 3, 2, 6, 9, 5),
+  x2 = c(7, 3, 8, 6, 6, 5, 6, 6, 4),
+  class = factor(c(1, 2, 2, 2, 1, 1, 1, 2, 2), levels = c(1, 2))
+)
 
-```{r r-script-before, echo = FALSE, fig.cap="작업 이전 R Markdown 파일", out.width='80%', fig.align='center'}
-knitr::include_graphics("images/r-script-before.png")
+group_mean <- function(.data, .group_var, .xvar) {...}
+pooled_variance <- function(.data, .group_var, .xvar) {...}
+fisher_ld <- function(.data, .group_var, .xvar) {...}
+
+w_hat <- fisher_ld(binaryclass2, class, c(x1, x2))
+print(w_hat)
 ```
 
-]
+---
 
---
+- R 스크립트를 목적별로 분류하여 각 목적에 맞는 폴더에 저장
+  + **데이터 생성: data-raw/ 폴더에 .R 파일로 저장**
+  + 함수 정의: R/ 폴더에 .R 파일로 저장
+  + 데이터 분석: vignettes/ 폴더에 .Rmd 파일로 저장
 
-.pull-right[
+```r
+*binaryclass2 <- tibble::tibble(
+* id = c(1:9),
+* x1 = c(5, 4, 7, 8, 3, 2, 6, 9, 5),
+* x2 = c(7, 3, 8, 6, 6, 5, 6, 6, 4),
+* class = factor(c(1, 2, 2, 2, 1, 1, 1, 2, 2), levels = c(1, 2))
+*)
 
-```{r r-script-after-data-raw, echo = FALSE, out.width='70%', fig.cap="데이터 생성: data-raw/ 폴더", fig.align='center'}
-knitr::include_graphics("images/r-after-data-raw.png")
+group_mean <- function(.data, .group_var, .xvar) {...}
+pooled_variance <- function(.data, .group_var, .xvar) {...}
+fisher_ld <- function(.data, .group_var, .xvar) {...}
+
+w_hat <- fisher_ld(binaryclass2, class, c(x1, x2))
+print(w_hat)
 ```
 
-```{r r-script-after-function-r, echo = FALSE, out.width='70%', fig.cap="함수 정의: R/ 폴더", fig.align='center'}
-knitr::include_graphics("images/r-after-function-r.png")
+---
+
+- R 스크립트를 목적별로 분류하여 각 목적에 맞는 폴더에 저장
+  + 데이터 생성: data-raw/ 폴더에 .R 파일로 저장
+  + **함수 정의: R/ 폴더에 .R 파일로 저장**
+  + 데이터 분석: vignettes/ 폴더에 .Rmd 파일로 저장
+
+```r
+binaryclass2 <- tibble::tibble(
+  id = c(1:9),
+  x1 = c(5, 4, 7, 8, 3, 2, 6, 9, 5),
+  x2 = c(7, 3, 8, 6, 6, 5, 6, 6, 4),
+  class = factor(c(1, 2, 2, 2, 1, 1, 1, 2, 2), levels = c(1, 2))
+)
+
+*group_mean <- function(.data, .group_var, .xvar) {...}
+*pooled_variance <- function(.data, .group_var, .xvar) {...}
+*fisher_ld <- function(.data, .group_var, .xvar) {...}
+
+w_hat <- fisher_ld(binaryclass2, class, c(x1, x2))
+print(w_hat)
 ```
 
-```{r r-script-after-vignettes, echo = FALSE, out.width='70%', fig.cap="데이터 분석 문서: vignettes/ 폴더", fig.align='center'}
-knitr::include_graphics("images/r-after-vignettes.png")
+---
+
+- R 스크립트를 목적별로 분류하여 각 목적에 맞는 폴더에 저장
+  + 데이터 생성: data-raw/ 폴더에 .R 파일로 저장
+  + 함수 정의: R/ 폴더에 .R 파일로 저장
+  + **데이터 분석: vignettes/ 폴더에 .Rmd 파일로 저장**
+
+```r
+binaryclass2 <- tibble::tibble(
+  id = c(1:9),
+  x1 = c(5, 4, 7, 8, 3, 2, 6, 9, 5),
+  x2 = c(7, 3, 8, 6, 6, 5, 6, 6, 4),
+  class = factor(c(1, 2, 2, 2, 1, 1, 1, 2, 2), levels = c(1, 2))
+)
+
+group_mean <- function(.data, .group_var, .xvar) {...}
+pooled_variance <- function(.data, .group_var, .xvar) {...}
+fisher_ld <- function(.data, .group_var, .xvar) {...}
+
+*w_hat <- fisher_ld(binaryclass2, class, c(x1, x2))
+*print(w_hat)
 ```
 
-]
-
 ---
+layout: false
 
@@
 
-  + **vignettes/**: 데이터 분석 문서 (.Rmd)
+  + **vignettes/**: 데이터 분석 문서 소스 (.Rmd)
+
+  + **doc/**: 배포용 데이터 분석 문서 (.html)
 

### 2021-09-15 draft.Rmd
수정 - 도움말 및 테스트 작성 스크린샷을 코드로 대체

@@
 
-# 함수 도움말 문서 작성
-
-- `{roxygen2}`를 이용한 함수 도움말 작성
-
-```{r r-function-manual, echo = FALSE, fig.cap="함수 도움말 작성 및 컴파일된 도움말 파일", out.width='49%', fig.align='center', fig.show='hold'}
-knitr::include_graphics(c(
-  "images/r-function-manual.png",
-  "images/r-function-manual-rendered.png"
-))
-```
-
----
-
-# 단위 테스트 작성
-
-- `{testthat}`을 이용한 예제 결과 재현성 테스트
+# `{roxygen2}`를 이용한 함수 도움말 문서 작성
 
@@
 
-```{r unit-test-book-example, echo = FALSE, fig.cap="단위 테스트 사용할 예제", out.width='50%', fig.align='center'}
-knitr::include_graphics("images/unit-test-book-example.png")
+```r
+#' 피셔 선형 판별 함수.
+#'
+#' 두 범주 데이터를 구분하는 피셔 선형 판별함수의 계수를 추정한다.
+#'
+#' @param .data 관측 데이터 프레임.
+#' @param .group_var 범주변수.
+#' @param .xvar 범주 분류에 사용될 변수.
+#' @return 선형 함수의 계수 벡터.
+#'
+#' @examples
+#' data(binaryclass2, package = "dmtr")
+#' fisher_ld(binaryclass2, class, c(x1, x2))
+#'
+#' @keywords discriminant-functions
+#' @export
+fisher_ld <- function(.data, .group_var, .xvar) {
+  ...
+}
+
 ```
@@
 
-```{r unit-test-script, echo = FALSE, out.width='70%', fig.cap="단위 테스트 작성: tests/ 폴더", fig.align='center'}
-knitr::include_graphics("images/unit-test-script.png")
-```
-
-```{r unit-test-result, echo = FALSE, out.width='50%', fig.cap="단위 테스트 결과", fig.align='center'}
-knitr::include_graphics("images/unit-test-result.png")
+```{r r-function-manual, echo = FALSE, fig.cap="함수 도움말 작성 및 컴파일된 도움말 파일 (.Rd)", out.width='90%', fig.align='center'}
+knitr::include_graphics("images/r-function-manual-fisher_ld.png")
 ```
@@
 ]
+
+---
+
+# `{testthat}`을 이용한 단위 테스트 작성
+
+- 예제 결과 재현성 테스트
+
+```{r lda-book-example-2, echo = FALSE, fig.cap="교재 예제", out.width='50%', fig.align='center'}
+knitr::include_graphics("images/lda-book-example-function.png")
+```
+
+--
+
+```{r lda-test, echo = TRUE}
+library(testthat); library(dmtr);
+test_that("Fisher discriminant function matches", {
+  local_edition(3)
+  expect_equal(
+    fisher_ld(binaryclass2, class, c(x1, x2)), #<<
+    c(x1 = -1.5080, x2 = 1.5418), #<<
+    tolerance = 1e-3,
+    ignore_attr = TRUE
+  )
+})
+```
 

### 2021-09-17 draft.Rmd
수정 - typo

@@
 
-```{r book, echo = FALSE, fig.cap="데이터미아닝 기법과 응용 / 한나래출판사 / 전치혁 저", out.height='400px', fig.align='center'}
+```{r book, echo = FALSE, fig.cap="데이터마이닝 기법과 응용 / 한나래출판사 / 전치혁 저", out.height='400px', fig.align='center'}
 knitr::include_graphics("images/book-cover.jpg")

### 2021-09-26 draft.Rmd
roxygen 코드 정리 - 스크롤 제거

@@
 
-```{r lda-mass-summary, echo = TRUE, out.width='50%', fig.cap="{MASS} 패키지 결과", fig.align='center'}
+```{r lda-mass-summary, echo = TRUE}
 library(MASS)
@@
 #'
-#' 두 범주 데이터를 구분하는 피셔 선형 판별함수의 계수를 추정한다.
+#' 두 범주 데이터를 구분하는 피셔 선형 판별함수의 
+#' 계수를 추정한다.
 #'
@@
 #' @export
-fisher_ld <- function(.data, .group_var, .xvar) {
-  ...
-}
-
+fisher_ld <- 
+  function(.data, .group_var, .xvar) {...}
 ```

### 2021-09-26 draft.Rmd
set knit hook option with {fansi} package to show colored text in code chunk output correctly

@@
       after_body: [css/rconf-logo.html]
-editor_options: 
-  chunk_output_type: console
 ---
@@
 ### `r rmarkdown::metadata$date`
+
+```{r, comment="", results="asis"}
+old.hooks <- fansi::set_knit_hooks(knitr::knit_hooks)
+```
 
@@
 
+  + **tests/**: 단위 테스트 (.R)
+
+---
+layout: true
+
+# 변화 관리
 ---
 
-# 변화 관리
+- 단위 테스트
 
-- 단위 테스트
   + 함수 구현의 수정, 패키지 업데이트 등의 변경사항 발생 후, 데이터 분석 결과가  옳게 나타나는지 확인하고, 오류 발생 시 빠르게 오류를 찾아 수정
+
   + 단위 테스트 생성 후, 데이터 분석 함수의 수정을 보다 적극적으로 자주 수행할 수 있어 리팩토링(refactoring) 생산성 증가
 
+---
+
 - 네임스페이스 명시화
-  + 함수 콜 수행 시 네임스페이스 충돌 방지
-      + `library(MASS); library(dplyr); tibble(x = 1:2, y = 2:3) %>% select(x)`
-      + `library(dplyr); library(MASS); tibble(x = 1:2, y = 2:3) %>% select(x)`
-      +  `dplyr::tibble(x = 1:2, y = 2:3) %>% dplyr::select(x)`
+
+  + 함수 콜 수행 시 prefix 사용, `package::function()`
+
+.pull-left[
+```{r}
+# detach(package:MASS)
+detach(package:dplyr)
+```
+
+```{r namespace-1, echo=TRUE, error=TRUE}
+library(MASS)
+library(dplyr)
+tibble(x = 1:2, y = 2:3) %>% 
+  select(x, y)
+```
+]
+
+--
+
+.pull-right[
+```{r}
+detach(package:MASS)
+detach(package:dplyr)
+```
+
+```{r namespace-2, echo=TRUE, error=TRUE}
+library(dplyr)
+library(MASS)
+tibble(x = 1:2, y = 2:3) %>% 
+  select(x, y)
+```
+]
 
 ---
+
+- 네임스페이스 명시화
+
+  + 함수 콜 수행 시 prefix 사용, `package::function()`
+
+```{r}
+detach(package:MASS)
+```
+
+```{r namespace-3, echo=TRUE, error=TRUE}
+dplyr::tibble(x = 1:2, y = 2:3) %>% 
+  dplyr::select(x, y)
+```
+
+---
+layout: false
 

### 2021-09-27 draft.Rmd
complete first draft

@@
   - 코드의 재사용성 향상
+
+3. 맺음말
 
@@
 library(MASS)
-MASS::lda(class ~ x1 + x2, data = binaryclass2)
+MASS::lda(class ~ x1 + x2, binaryclass2)
 ```
@@
 ]
-
---
 
@@
 ---
+
+# 코드 테스트
+
+.pull-left[
+- 리팩토링(refactoring) 생산성 증가
+
+- R 버전 및 패키지 업데이트에 따른 이상 유무 확인
+
+- One is better than nothing
+]
+
+.pull-right[
+
+```{r refactoring-book, echo = FALSE, fig.cap="Refactoring: improving the design of existing code / Addison-Wesley Professional / Martin Fowler", out.width='40%', fig.align='center'}
+knitr::include_graphics("images/refact2.jpeg")
+```
+
+]
+
+---
 layout: true
 
-# 변화 관리
----
-
-- 단위 테스트
-
-  + 함수 구현의 수정, 패키지 업데이트 등의 변경사항 발생 후, 데이터 분석 결과가  옳게 나타나는지 확인하고, 오류 발생 시 빠르게 오류를 찾아 수정
-
-  + 단위 테스트 생성 후, 데이터 분석 함수의 수정을 보다 적극적으로 자주 수행할 수 있어 리팩토링(refactoring) 생산성 증가
+# 코드 모호성 제거
 
 ---
+- 네임스페이스 명시화 / `package::function()`
 
-- 네임스페이스 명시화
+  + 패키지 로드 순서에 따라 결과가 달라지거나 오류 발생하는 문제
 
-  + 함수 콜 수행 시 prefix 사용, `package::function()`
+--
 
@@
 tibble(x = 1:2, y = 2:3) %>% 
-  select(x, y)
+  select(x)
 ```
@@
 ---
+- 네임스페이스 명시화 / `package::function()`
 
-- 네임스페이스 명시화
+  + 어떤 패키지의 함수를 사용할 것인지 명시화
 
-  + 함수 콜 수행 시 prefix 사용, `package::function()`
+--
 
@@
 dplyr::tibble(x = 1:2, y = 2:3) %>% 
-  dplyr::select(x, y)
+  dplyr::select(x)
 ```
+
+---
+- 네임스페이스 명시화 / `package::function()`
+
+  + 코드를 이해하기 쉽도록 도움
+
+--
+
+.pull-left[
+```{r, echo=TRUE, message=FALSE}
+library(dmtr)
+library(corrr)
+library(dplyr)
+binaryclass2 %>% 
+  select(x1, x2) %>% 
+  correlate()
+```
+]
+
+--
+
+.pull-right[
+```{r, echo=TRUE, message=FALSE}
+library(dmtr)
+library(dplyr)
+binaryclass2 %>% 
+  select(x1, x2) %>% 
+  corrr::correlate()
+```
+
+]
 
@@
 
--
+  + 하드코딩 지양
+
+  + **tidy eval** 프레임워크 활용
+
+```r
+fisher_ld <- function(.data, .group_var, .xvar) {
+* .group_var <- rlang::enquo(.group_var)
+* .xvar <- rlang::enquo(.xvar)
+* mu_hat <- group_mean(.data, !!.group_var, !!.xvar)
+
+  if (length(mu_hat) != 2L) {
+    stop("number of levels of .group_var must be 2.")
+  }
+
+* sigma_hat <- pooled_variance(.data, !!.group_var, !!.xvar)
+
+  res <- solve(sigma_hat) %*% purrr::reduce(mu_hat, `-`) %>% drop()
+  attr(res, "group") <- attr(mu_hat, "group")
+  return(res)
+}
+
+* fisher_ld(binaryclass2, class, c(x1, x2))
+```
 
 ---
+class: inverse, center, middle
+
+# 맺음말
+
+---
+
+# 데이터 분석가에게 패키지 개발이란?
+
+- 분석 과정 및 결과의 배포
+
+  + 데이터, 분석 코드, 리포트 등
+
+  + 재현가능성 검증
+
+- 생산성 향상을 위한 배움의 과정
+
+  + 프로젝트 관리
+
+  + 코드 테스트
+
+  + 코딩 스타일
+
+- 경험의 축적이 중요
+
+  + 쉬운 목표 설정
+
+  + 기존에 수행한 가장 간단한 분석 프로젝트로 시작
+
+  + 하나의 함수부터 시작

### 2021-10-08 draft.Rmd
구성 및 컨텐츠 변경 - "배경"

@@
 
+  - 동기
+
   - 작업 과정
@@
 ---
-
-# `{bookdown}`을 이용한 R 예제 책 만들기
+class: middle
+layout: true
 
@@
 
+---
+
+---
+
 .pull-right[
+
+- 데이터 마이닝 알고리즘을 잘 이해할 수 있게 깔끔하게 구성된 교재
+
+]
+
+---
+
+.pull-right[
+
+- 데이터 마이닝 알고리즘을 잘 이해할 수 있게 깔끔하게 구성된 교재
+
+- 단순히 기존 패키지를 사용하는 것보다 깊이 있는 내용을 제공
+
+]
+
+---
+
+.pull-right[
+
+- 데이터 마이닝 알고리즘을 잘 이해할 수 있게 깔끔하게 구성된 교재
+
+- 단순히 기존 패키지를 사용하는 것보다 깊이 있는 내용을 제공
+
+- 간단하면서 결과를 눈으로 확인할 수 있는 예제들
+
+]
+
+---
+
+.pull-right[
+
+- 데이터 마이닝 알고리즘을 잘 이해할 수 있게 깔끔하게 구성된 교재
+
+- 단순히 기존 패키지를 사용하는 것보다 깊이 있는 내용을 제공
+
+- 간단하면서 결과를 눈으로 확인할 수 있는 예제들
+
+- **.red[예제 결과 도출을 위한 프로그램 구현은 제공되지 않음]**
+
+]
+
+---
+class: middle
+layout: false
+
+.pull-left[
+
+```{r rlogo, out.width="70%"}
+knitr::include_graphics(c("images/Rlogo.png"))
+```
+
+]
+
+.pull-right[
+
+```{r bookdownlogo, out.width="70%"}
+knitr::include_graphics(c("images/bookdown-logo.png"))
+```
+
+]
+
+---
+class: middle
+layout: true
+
+.pull-left[
 
@@
 ---
+
+---
+
+.pull-right[
+
+- 교재를 온라인 버전으로 전환
+
+]
+
+---
+
+.pull-right[
+
+- 온라인 버전 보조 교재로 활용
+
+- 예제를 R로 구현한 내용을 추가
+
+]
+
+---
+class: inverse, center, middle
+layout: false
+
+# 데이터 분석 문서화에서 패키지 개발로 전환
+
+## 동기
+
+---
+layout: false
 

### 2021-10-09 draft.Rmd
패키지 작성 동기 구성 수정

@@
 
-- 데이터 마이닝 알고리즘을 잘 이해할 수 있게 깔끔하게 구성된 교재
-
-]
-
----
-
-.pull-right[
-
-- 데이터 마이닝 알고리즘을 잘 이해할 수 있게 깔끔하게 구성된 교재
-
-- 단순히 기존 패키지를 사용하는 것보다 깊이 있는 내용을 제공
-
-]
-
----
-
-.pull-right[
-
-- 데이터 마이닝 알고리즘을 잘 이해할 수 있게 깔끔하게 구성된 교재
-
-- 단순히 기존 패키지를 사용하는 것보다 깊이 있는 내용을 제공
-
-- 간단하면서 결과를 눈으로 확인할 수 있는 예제들
-
-]
-
----
-
-.pull-right[
-
-- 데이터 마이닝 알고리즘을 잘 이해할 수 있게 깔끔하게 구성된 교재
-
-- 단순히 기존 패키지를 사용하는 것보다 깊이 있는 내용을 제공
-
-- 간단하면서 결과를 눈으로 확인할 수 있는 예제들
-
-- **.red[예제 결과 도출을 위한 프로그램 구현은 제공되지 않음]**
-
-]
-
----
-class: middle
-layout: false
-
-.pull-left[
-
-```{r rlogo, out.width="70%"}
-knitr::include_graphics(c("images/Rlogo.png"))
-```
-
-]
-
-.pull-right[
-
-```{r bookdownlogo, out.width="70%"}
-knitr::include_graphics(c("images/bookdown-logo.png"))
-```
-
-]
-
----
-class: middle
-layout: true
-
-.pull-left[
-
-```{r online-book-webshot, include=FALSE, cache=TRUE}
-webshot::webshot(
-  url = "https://youngroklee-ml.github.io/data-mining-book/index.html",
-  file = "images/onlinebook-main.jpg",
-  zoom = 3
-)
-```
-
-```{r online-book, echo = FALSE, fig.cap="데이터마이닝 with R / 전치혁, 이혜선, 이종석, 이영록 / http://youngroklee-ml.github.io/data-mining-book/", out.height='400px', fig.align='center'}
-knitr::include_graphics("images/onlinebook-main.jpg")
-```
-
-]
-
----
-
----
-
-.pull-right[
-
-- 교재를 온라인 버전으로 전환
-
-]
-
----
-
-.pull-right[
-
-- 온라인 버전 보조 교재로 활용
-
-- 예제를 R로 구현한 내용을 추가
-
-]
-
----
-class: inverse, center, middle
-layout: false
-
-# 데이터 분석 문서화에서 패키지 개발로 전환
-
-## 동기
-
----
-layout: false
-
-# 분석 예제: 선형 분류분석
-
-.pull-left[
-
-```{r lda-example, echo=FALSE, out.width='49%', fig.align='center', fig.show='hold'}
+```{r lda-example-plot, echo=FALSE, out.width='80%', fig.align='center', fig.cap="예제: 선형 분류분석"}
 data(binaryclass2, package = "dmtr")
 
-binaryclass2 %>% 
-  gt::gt()
-```
-
-]
-
-.pull-right[
-
-```{r lda-example-plot, echo=FALSE, out.width='90%', fig.align='center'}
 binaryclass2 %>% 
@@
 ---
-layout: true
 
-# 교재용 R 패키지 작성의 필요성
+.pull-right[
 
-- 알고리즘의 단계별 결과를 교재 내용대로 보여주기 위해 함수를 세분화할 필요
+- 데이터 마이닝 알고리즘을 잘 이해할 수 있게 깔끔하게 구성된 교재
+{{content}}
+
+]
+
+--
+
+- 단순히 기존 패키지를 사용하는 것보다 깊이 있는 내용을 제공
+{{content}}
+
+--
+
+- 간단하면서 결과를 눈으로 확인할 수 있는 예제들
+{{content}}
+
+--
+
+- **예제 결과 도출을 위한 프로그램 구현은 제공되지 않음**
+{{content}}
+
+---
+class: middle
+layout: false
 
@@
 
-```{r book-covarince, echo = FALSE, fig.cap="교재 예제", out.width='80%', fig.align='center'}
-knitr::include_graphics("images/book-covariance.png")
+```{r rlogo, out.width="70%"}
+knitr::include_graphics(c("images/Rlogo.png"))
+```
+
+]
+
+.pull-right[
+
+```{r bookdownlogo, out.width="70%"}
+knitr::include_graphics(c("images/bookdown-logo.png"))
 ```
@@
 ---
+class: middle
+
+.pull-left[
+
+```{r online-book-webshot, include=FALSE, cache=TRUE}
+webshot::webshot(
+  url = "https://youngroklee-ml.github.io/data-mining-book/index.html",
+  file = "images/onlinebook-main.jpg",
+  zoom = 3
+)
+```
+
+```{r online-book, echo = FALSE, fig.cap="데이터마이닝 with R / 전치혁, 이혜선, 이종석, 이영록 / http://youngroklee-ml.github.io/data-mining-book/", out.height='400px', fig.align='center'}
+knitr::include_graphics("images/onlinebook-main.jpg")
+```
+
+]
+
+.pull-right[
+  {{content}}
+]
+
+--
+
+- 온라인 버전 보조 교재로 활용
+{{content}}
+
+--
+
+- 예제를 R로 구현한 내용을 추가
+{{content}}
+
+--
+
+- 기존 R 패키지가 제공하지 않는 세부 결과 도출과정을 구현
+{{content}}
+
+--
+
+- 기존 R 패키지가 교재 설명과는 다른 알고리즘 사용한 경우에 대한 추가 설명
+{{content}}
+
+---
+class: middle
+
+```{r bookdown-folder, out.width="50%"}
+knitr::include_graphics("images/bookdown-folder.png")
+```
+
+---
+class: inverse, center, middle
+layout: false
+
+# 데이터 분석 문서화에서 패키지 개발로 전환
+
+## 동기
+
+---
+class: middle, center
+
+# 사용자 함수의 난립
+
+---
+class: middle
+
+.left-column[
+
+```{r rmd-single-image, out.width="80%"}
+knitr::include_graphics("images/rmd_image.png")
+```
+
+]
 
@@
 
-```{r lda-mass-summary, echo = TRUE}
-library(MASS)
-MASS::lda(class ~ x1 + x2, binaryclass2)
+{{content}}
+
+]
+
+--
+
+데이터 생성
+{{content}}
+
+--
+
+**함수 1**
+{{content}}
+
+--
+
+**함수 2**
+{{content}}
+
+--
+
+분석
+{{content}}
+
+--
+
+**함수 3**
+{{content}}
+
+--
+
+분석
+{{content}}
+
+--
+
+데이터 생성
+{{content}}
+
+--
+
+**함수 4**
+{{content}}
+
+--
+
+**함수 5**
+{{content}}
+
+--
+
+데이터 분석
+{{content}}
+
+--
+
+**...**
+{{content}}
+
+---
+class: middle
+
+.pull-left[
+
+```{r multiple-rmd-image, out.width="80%"}
+knitr::include_graphics("images/bookdown-folder.png")
+```
+
+]
+
+--
+
+.pull-right[
+
+```{r disorganized-spoons, out.width="90%"}
+knitr::include_graphics("images/rick-mason-2FaCKyEEtis-unsplash.jpg")
 ```
@@
 ---
+class: middle, center
+
+# 지금은 맞고, 나중엔 틀릴수도?
+
+---
+class: middle
+
+.pull-left[
+
+```{r namespace-1, echo=TRUE, error=TRUE}
+library(dplyr)
+
+data(binaryclass2, package = "dmtr")
+
+binaryclass2 %>% 
+  select(x1, x2) #<<
+```
+
+]
+
+--
 
 .pull-right[
+```{r}
+detach(package:dplyr)
+```
 
-```{r lda-dmtr-group-variance, echo = TRUE}
-library(dmtr)
-group_variance(binaryclass2, class, x1:x2)
+```{r namespace-2, echo=TRUE, error=TRUE}
+library(dplyr)
+
+data(binaryclass2, package = "dmtr")
+
+library(MASS)
+fit <- lda(class ~ x1 + x2, binaryclass2)
+
+binaryclass2 %>% 
+  select(x1, x2) #<<
+```
+
+```{r}
+detach(package:MASS)
 ```
@@
 ---
+class: middle
+
+.pull-left[
+
+**`{tidyr}` 0.x**
+
+```r
+df <- tibble(
+ a = list(c("a", "b"), "c"),
+ b = list(1:2, 3),
+ c = c(11, 22)
+)
+* df %>% tidyr::unnest(a, b)
+```
+
+```{r unnest-old}
+df <- tibble(
+ a = list(c("a", "b"), "c"),
+ b = list(1:2, 3),
+ c = c(11, 22)
+)
+df %>% tidyr::unnest_legacy(a, b)
+```
+
+]
+
+--
 
@@
 
-```{r lda-dmtr-pooled-variance, echo = TRUE}
-library(dmtr)
-pooled_variance(binaryclass2, class, x1:x2)
+**`{tidyr}` 1.0.0**
+
+```{r unnest-new, echo=TRUE}
+df <- tibble(
+ a = list(c("a", "b"), "c"),
+ b = list(1:2, 3),
+ c = c(11, 22)
+)
+df %>% tidyr::unnest(c(a, b)) #<<
 ```
@@
 ---
-layout: true
+class: middle
 
-# 교재용 R 패키지 작성의 필요성
+- 오류 발생 시 수정에 많은 시간 소요
 
-- 기존 R 패키지 수행 결과가 교재의 예제와 정확히 일치하지 않는 부분에 대한 혼동을 줄일 필요
-
-```{r lda-book-example, echo = FALSE, fig.cap="교재 예제", out.width='50%', fig.align='center'}
-knitr::include_graphics("images/lda-book-example-function.png")
-```
-
----
-
---
-
-```{r lda-mass-result, echo = TRUE, out.width='50%', fig.cap="{MASS} 패키지 결과", fig.align='center'}
-fit <- MASS::lda(class ~ x1 + x2, data = binaryclass2)
-coef(fit) %>% drop()
-```
-
---
-
-```{r lda-mass-result-transform, echo = TRUE, out.width='50%', fig.cap="{MASS} 패키지 결과 변환", fig.align='center'}
-scaling_adjust <- 2.201359; sign_adjust <- -1;
-(coef(fit) * scaling_adjust * sign_adjust) %>% drop()
-```
-
----
-
-```{r lda-dmtr-result, echo = TRUE}
-dmtr::fisher_ld(binaryclass2, class, x1:x2)
-```
-
----
-layout: false
-
-# 교재용 R 패키지 작성의 필요성
-
-- 분석 알고리즘을 교재와 일치하도록 함수로 구현
-
-```r
-fisher_ld <- function(.data, .group_var, .xvar) {
-  .group_var <- rlang::enquo(.group_var)
-  .xvar <- rlang::enquo(.xvar)
-
-* mu_hat <- group_mean(.data, !!.group_var, !!.xvar)
-
-  if (length(mu_hat) != 2L) {
-    stop("number of levels of .group_var must be 2.")
-  }
-
-* sigma_hat <- pooled_variance(.data, !!.group_var, !!.xvar)
-
-* res <- solve(sigma_hat) %*% purrr::reduce(mu_hat, `-`) %>% drop()
-
-  attr(res, "group") <- attr(mu_hat, "group")
-
-  return(res)
-}
-```
-
-```{r, include=FALSE}
-detach("package:MASS", character.only = TRUE)
-```
+- 코드 실행 오류는 없으나 결과가 틀린 경우는 오류 발견 자체가 어려움
 
@@
 
---
-
-.pull-left[
-```{r}
-# detach(package:MASS)
-detach(package:dplyr)
-```
-
-```{r namespace-1, echo=TRUE, error=TRUE}
-library(MASS)
-library(dplyr)
-tibble(x = 1:2, y = 2:3) %>% 
-  select(x)
-```
-]
-
---
-
-.pull-right[
-```{r}
-detach(package:MASS)
-detach(package:dplyr)
-```
-
-```{r namespace-2, echo=TRUE, error=TRUE}
-library(dplyr)
-library(MASS)
-tibble(x = 1:2, y = 2:3) %>% 
-  select(x)
-```
-]
-
 ---
@@
 --
-
-```{r}
-detach(package:MASS)
-```
 

### 2021-10-10 draft.Rmd
작업 과정 내용 구성 변경

@@
 
-## 배경
+---
+class: inverse, center, middle
+
+# 배경
 
@@
 
-# 데이터 분석 문서화에서 패키지 개발로 전환
-
-## 동기
+# 동기
 
@@
 ---
-class: middle
+# 사용자 함수의 난립
 
-.left-column[
+.pull-left[
 
@@
 ---
-class: middle
+
+# 사용자 함수의 난립
 
@@
 
-# 데이터 분석 문서화에서 패키지 개발로 전환
-
-## 작업 과정
+# 작업 과정
 
 ---
-
-# 현실적인 작업범위 및 기대수준 정의
-
-- 참고교재
-
-- 실무용 아님
-
-- 제한된 유연성
-
----
-
-# 편리한 작업환경 선택
+class: center, middle
 
@@
 
-- RStudio IDE
-
-- `{devtools}`, `{usethis}`, `{testthat}`, etc.
-
-- Git and GitHub
+```{r r-packages-book, echo = FALSE, fig.cap="R Packages / O'Reilly / Hadley Wickham", out.width='80%', fig.align='center'}
+knitr::include_graphics("images/r-pkgs-cover.png")
+```
 
@@
 
-```{r r-packages-book, echo = FALSE, fig.cap="R Packages / O'Reilly / Hadley Wickham", out.width='40%', fig.align='center'}
-knitr::include_graphics("images/r-pkgs-cover.png")
+```{r r-devtools-logo, echo = FALSE, out.width='80%', fig.align='center'}
+knitr::include_graphics("images/devtools-logo.svg")
 ```
@@
 ---
-layout: true
+class: center, middle
 
-# R 스크립트 분류 및 재구성
+# R Markdown 문서 해체 및 재구성
 
@@
 
-- R 스크립트를 목적별로 분류하여 각 목적에 맞는 폴더에 저장
-  + 데이터 생성: data-raw/ 폴더에 .R 파일로 저장
-  + 함수 정의: R/ 폴더에 .R 파일로 저장
-  + 데이터 분석: vignettes/ 폴더에 .Rmd 파일로 저장
+# R Markdown 문서 해체 및 재구성
 
-```r
-binaryclass2 <- tibble::tibble(
-  id = c(1:9),
-  x1 = c(5, 4, 7, 8, 3, 2, 6, 9, 5),
-  x2 = c(7, 3, 8, 6, 6, 5, 6, 6, 4),
-  class = factor(c(1, 2, 2, 2, 1, 1, 1, 2, 2), levels = c(1, 2))
+.pull-left[
+
+````
+```{r}`r ''`
+binaryclass2 <- tribble(
+  ~id, ~x1, ~x2, ~class,
+  1, 5, 7, 1,
+  2, 4, 3, 2,
+  3, 7, 8, 2,
+  4, 8, 6, 2,
+  5, 3, 6, 1,
+  6, 2, 5, 1,
+  7, 6, 6, 1,
+  8, 9, 6, 2,
+  9, 5, 4, 2
 )
 
-group_mean <- function(.data, .group_var, .xvar) {...}
-pooled_variance <- function(.data, .group_var, .xvar) {...}
-fisher_ld <- function(.data, .group_var, .xvar) {...}
+group_mean <- function(...) {...}
+pooled_variance <- function(...) {...}
+fisher_ld <- function(...) {...}
 
-w_hat <- fisher_ld(binaryclass2, class, c(x1, x2))
+w_hat <- fisher_ld(
+  binaryclass2, class, x1:x2)
 print(w_hat)
+```
+````
+
+]
+
+---
+
+# R Markdown 문서 해체 및 재구성
+
+.pull-left[
+
+````
+```{r}`r ''`
+*binaryclass2 <- tribble(
+*  ~id, ~x1, ~x2, ~class,
+*  1, 5, 7, 1,
+*  2, 4, 3, 2,
+*  3, 7, 8, 2,
+*  4, 8, 6, 2,
+*  5, 3, 6, 1,
+*  6, 2, 5, 1,
+*  7, 6, 6, 1,
+*  8, 9, 6, 2,
+*  9, 5, 4, 2
+*)
+
+group_mean <- function(...) {...}
+pooled_variance <- function(...) {...}
+fisher_ld <- function(...) {...}
+
+w_hat <- fisher_ld(
+  binaryclass2, class, x1:x2)
+print(w_hat)
+```
+````
+
+]
+
+.pull-right[
+
+```{r data-raw-example, out.width='30%', fig.align='center'}
+knitr::include_graphics(c(
+  "images/data-raw-folder-icon.png", 
+  "images/data-raw-binaryclass.png"))
+```
+
+]
+
+---
+
+# R Markdown 문서 해체 및 재구성
+
+.pull-left[
+
+````
+```{r}`r ''`
+binaryclass2 <- tribble(
+  ~id, ~x1, ~x2, ~class,
+  1, 5, 7, 1,
+  2, 4, 3, 2,
+  3, 7, 8, 2,
+  4, 8, 6, 2,
+  5, 3, 6, 1,
+  6, 2, 5, 1,
+  7, 6, 6, 1,
+  8, 9, 6, 2,
+  9, 5, 4, 2
+)
+
+*group_mean <- function(...) {...}
+*pooled_variance <- function(...) {...}
+*fisher_ld <- function(...) {...}
+
+w_hat <- fisher_ld(
+  binaryclass2, class, x1:x2)
+print(w_hat)
+```
+````
+
+]
+
+.pull-right[
+
+```{r r-example, out.width='30%', fig.align='center'}
+knitr::include_graphics(c(
+  "images/r-folder-icon.png", 
+  "images/r-within-group.png",
+  "images/r-discriminant-functions.png"))
+```
+
+]
+
+---
+
+# R Markdown 문서 해체 및 재구성
+
+.pull-left[
+
+````
+```{r}`r ''`
+binaryclass2 <- tribble(
+  ~id, ~x1, ~x2, ~class,
+  1, 5, 7, 1,
+  2, 4, 3, 2,
+  3, 7, 8, 2,
+  4, 8, 6, 2,
+  5, 3, 6, 1,
+  6, 2, 5, 1,
+  7, 6, 6, 1,
+  8, 9, 6, 2,
+  9, 5, 4, 2
+)
+
+group_mean <- function(...) {...}
+pooled_variance <- function(...) {...}
+fisher_ld <- function(...) {...}
+
+*w_hat <- fisher_ld(
+*  binaryclass2, class, x1:x2)
+*print(w_hat)
+```
+````
+
+]
+
+.pull-right[
+
+```{r vignettes-example, out.width='30%', fig.align='center'}
+knitr::include_graphics(c(
+  "images/vignettes-folder-icon.png", 
+  "images/vignettes-discriminant-analysis.png"))
+```
+
+]
+
+---
+
+# R Markdown 문서 해체 및 재구성
+
+.pull-left[
+
+````
+```{r}`r ''`
+library(dmtr)
+
+w_hat <- fisher_ld(
+  binaryclass2, class, x1:x2)
+print(w_hat)
+```
+````
+
+]
+
+.pull-right[
+
+```{r vignettes-example-2, out.width='30%', fig.align='center'}
+knitr::include_graphics(c(
+  "images/vignettes-folder-icon.png", 
+  "images/vignettes-discriminant-analysis.png"))
+```
+
+]
+
+---
+class: middle, center
+
+# 함수 도움말 문서 작성
+
+---
+
+# 함수 도움말 문서 작성
+
+```{r roxygen-logo, out.width='30%'}
+knitr::include_graphics(
+  "images/roxygen2-logo.png"
+)
 ```
@@
 ---
+layout: true
 
-- R 스크립트를 목적별로 분류하여 각 목적에 맞는 폴더에 저장
-  + **데이터 생성: data-raw/ 폴더에 .R 파일로 저장**
-  + 함수 정의: R/ 폴더에 .R 파일로 저장
-  + 데이터 분석: vignettes/ 폴더에 .Rmd 파일로 저장
-
-```r
-*binaryclass2 <- tibble::tibble(
-* id = c(1:9),
-* x1 = c(5, 4, 7, 8, 3, 2, 6, 9, 5),
-* x2 = c(7, 3, 8, 6, 6, 5, 6, 6, 4),
-* class = factor(c(1, 2, 2, 2, 1, 1, 1, 2, 2), levels = c(1, 2))
-*)
-
-group_mean <- function(.data, .group_var, .xvar) {...}
-pooled_variance <- function(.data, .group_var, .xvar) {...}
-fisher_ld <- function(.data, .group_var, .xvar) {...}
-
-w_hat <- fisher_ld(binaryclass2, class, c(x1, x2))
-print(w_hat)
-```
-
----
-
-- R 스크립트를 목적별로 분류하여 각 목적에 맞는 폴더에 저장
-  + 데이터 생성: data-raw/ 폴더에 .R 파일로 저장
-  + **함수 정의: R/ 폴더에 .R 파일로 저장**
-  + 데이터 분석: vignettes/ 폴더에 .Rmd 파일로 저장
-
-```r
-binaryclass2 <- tibble::tibble(
-  id = c(1:9),
-  x1 = c(5, 4, 7, 8, 3, 2, 6, 9, 5),
-  x2 = c(7, 3, 8, 6, 6, 5, 6, 6, 4),
-  class = factor(c(1, 2, 2, 2, 1, 1, 1, 2, 2), levels = c(1, 2))
-)
-
-*group_mean <- function(.data, .group_var, .xvar) {...}
-*pooled_variance <- function(.data, .group_var, .xvar) {...}
-*fisher_ld <- function(.data, .group_var, .xvar) {...}
-
-w_hat <- fisher_ld(binaryclass2, class, c(x1, x2))
-print(w_hat)
-```
-
----
-
-- R 스크립트를 목적별로 분류하여 각 목적에 맞는 폴더에 저장
-  + 데이터 생성: data-raw/ 폴더에 .R 파일로 저장
-  + 함수 정의: R/ 폴더에 .R 파일로 저장
-  + **데이터 분석: vignettes/ 폴더에 .Rmd 파일로 저장**
-
-```r
-binaryclass2 <- tibble::tibble(
-  id = c(1:9),
-  x1 = c(5, 4, 7, 8, 3, 2, 6, 9, 5),
-  x2 = c(7, 3, 8, 6, 6, 5, 6, 6, 4),
-  class = factor(c(1, 2, 2, 2, 1, 1, 1, 2, 2), levels = c(1, 2))
-)
-
-group_mean <- function(.data, .group_var, .xvar) {...}
-pooled_variance <- function(.data, .group_var, .xvar) {...}
-fisher_ld <- function(.data, .group_var, .xvar) {...}
-
-*w_hat <- fisher_ld(binaryclass2, class, c(x1, x2))
-*print(w_hat)
-```
-
----
-layout: false
-
-# `{roxygen2}`를 이용한 함수 도움말 문서 작성
+# 함수 도움말 문서 작성
 
@@
 
---
+---
+
+---
+
+.pull-right[
+
+```{r man-example, out.width='30%', fig.align='center'}
+knitr::include_graphics(c(
+  "images/man-folder-icon.png", 
+  "images/man-fisher_ld.png"))
+```
+
+]
+
+---
 
@@
 ---
+class: middle, center
+layout: false
 
-# `{testthat}`을 이용한 단위 테스트 작성
+# 코드 테스트
 
-- 예제 결과 재현성 테스트
+---
 
-```{r lda-book-example-2, echo = FALSE, fig.cap="교재 예제", out.width='50%', fig.align='center'}
+# 코드 테스트
+
+```{r testthat-logo, out.width='30%'}
+knitr::include_graphics(
+  "images/testthat-logo.png"
+)
+```
+
+---
+layout: true
+
+# 코드 테스트
+
+예제 결과 재현성 테스트
+
+```{r lda-book-example-2, echo = FALSE, out.width='50%', fig.align='center'}
 knitr::include_graphics("images/lda-book-example-function.png")
@@
 
---
+---
+
+---
+
+```{r tests-example, out.width='10%', fig.align='center'}
+knitr::include_graphics(c(
+  "images/tests-folder-icon.png", 
+  "images/tests-discriminant-functions.png"))
+```
+
+---
 
@@
   expect_equal(
-    fisher_ld(binaryclass2, class, c(x1, x2)), #<<
+    fisher_ld(binaryclass2, class, x1:x2), #<<
     c(x1 = -1.5080, x2 = 1.5418), #<<
@@
 ---
+class: middle, center
+layout: false
 
-# 패키지 사이트 생성
+# 패키지 웹사이트 생성
 
-- `{pkgdown}`을 이용한 패키지 웹사이트 생성
+---
 
-```{r pkgdown-site, echo = FALSE, out.width='50%', fig.cap="패키지 웹사이트", fig.align='center'}
+# 패키지 웹사이트 생성
+
+```{r pkgdown-logo, out.width='30%'}
+knitr::include_graphics(
+  "images/pkgdown-logo.png"
+)
+```
+
+---
+
+# 패키지 웹사이트 생성
+
+```{r pkgdown-site, echo = FALSE, out.width='70%', fig.cap="https://youngroklee-ml.github.io/dmtr/", fig.align='center'}
 knitr::include_graphics("images/pkgdown-site.gif")
@@
   + 하나의 함수부터 시작
+
+---
+
+# 현실적인 작업범위 및 기대수준 정의
+
+- 참고교재
+
+- 실무용 아님
+
+- 제한된 유연성
+
+---
+
+# 편리한 작업환경 선택
+
+.pull-left[
+
+- RStudio IDE
+
+- `{devtools}`, `{usethis}`, `{testthat}`, etc.
+
+- Git and GitHub
+
+]

### 2021-10-10 draft.Rmd
배운 점 재구성

@@
 
-1. 데이터 분석 문서화에서 패키지 개발로 전환
-
-  - 배경
-
-  - 동기
-
-  - 작업 과정
-
-2. 데이터 분석가에게 유용한 개발자 관점
-
-  - 프로젝트 관리
-
-  - 변화 관리
-
-  - 코드의 재사용성 향상
-
-3. 맺음말
+  - 패키지 개발 배경
+
+  - 주요 작업 과정
+
+  - 데이터 분석가가 배우면 유용한 개발자 관점
+
+  - 맺음말
 
@@
 
-# 데이터 분석 문서화에서 패키지 개발로 전환
-
----
-class: inverse, center, middle
-
-# 배경
-
----
-class: middle
+# 패키지 개발 배경
+
+---
+class: center, middle
+
+# 데이터마이닝 교재 참고자료 제작
+
+---
 layout: true
+
+# 데이터마이닝 교재 참고자료 제작
 
@@
 
-- 데이터 마이닝 알고리즘을 잘 이해할 수 있게 깔끔하게 구성된 교재
-{{content}}
-
-]
-
---
-
-- 단순히 기존 패키지를 사용하는 것보다 깊이 있는 내용을 제공
-{{content}}
+- 데이터 마이닝 이론과 알고리즘을 잘 이해할 수 있게 깔끔하게 구성된 교재
+{{content}}
+
+]
 
@@
 ---
-class: middle
 layout: false
+
+# 데이터마이닝 교재 참고자료 제작
 
@@
 ---
-class: middle
+# 데이터마이닝 교재 참고자료 제작
 
@@
 
-- 기존 R 패키지가 제공하지 않는 세부 결과 도출과정을 구현
-{{content}}
-
---
-
-- 기존 R 패키지가 교재 설명과는 다른 알고리즘 사용한 경우에 대한 추가 설명
-{{content}}
-
----
-class: middle
+- 단순히 기존 R 패키지를 사용하는 것보다 깊이 있는 내용을 제공
+{{content}}
+
+---
+
+# 데이터마이닝 교재 참고자료 제작
 
@@
 ---
-class: inverse, center, middle
-layout: false
-
-# 동기
-
----
 class: middle, center
@@
 
-```{r disorganized-spoons, out.width="90%"}
-knitr::include_graphics("images/rick-mason-2FaCKyEEtis-unsplash.jpg")
+```{r messy-photo, out.width="90%"}
+knitr::include_graphics("images/xavi-cabrera-kn-UmDZQDjM-unsplash.jpg")
 ```
@@
 
-# 지금은 맞고, 나중엔 틀릴수도?
-
----
-class: middle
+# 지금은 맞지만, 나중엔 틀릴수도?
+
+---
+# 지금은 맞지만, 나중엔 틀릴수도?
 
@@
 ---
-class: middle
+# 지금은 맞지만, 나중엔 틀릴수도?
 
@@
 ---
-class: middle
+
+# 지금은 맞지만, 나중엔 틀릴수도?
 
@@
 
-# 작업 과정
+# 주요 작업 과정
 
@@
 
-# 데이터 분석가에게 유용한 개발자 관점
-
----
-
-# 프로젝트 관리
+# 데이터 분석가가 배우면 유용한 개발자 관점
+
+---
+class: center, middle
+
+# 체계적인 프로젝트 폴더 구조
+
+---
+
+# 체계적인 프로젝트 폴더 구조
+
+- 프로젝트 폴더
+
+```{r package-folder-structure, out.width='70%'}
+knitr::include_graphics("images/package-folder-structure.png")
+```
+
+---
+
+# 체계적인 프로젝트 폴더 구조
+
+- **data-raw/**: 데이터 생성/쿼리 R 스크립트 (.R)
+
+```{r data-raw-folder-files, out.width='70%'}
+knitr::include_graphics("images/data-raw-folder-files.png")
+```
+
+---
+
+# 체계적인 프로젝트 폴더 구조
+
+- **data/**: 생성된 데이터 파일 (.rda)
+
+```{r data-folder-files, out.width='50%'}
+knitr::include_graphics("images/data-folder-files.png")
+```
+
+---
+
+# 체계적인 프로젝트 폴더 구조
+
+- **R/**: 재사용을 위해 모듈화된 데이터 분석 함수 (.R)
+
+```{r r-folder-files, out.width='70%'}
+knitr::include_graphics("images/r-folder-files.png")
+```
+
+---
+
+# 체계적인 프로젝트 폴더 구조
+
+- **vignettes/**: 데이터 분석 문서 소스 (.Rmd)
+
+```{r vignettes-folder-files, out.width='70%'}
+knitr::include_graphics("images/vignettes-folder-files.png")
+```
+
+---
+
+# 체계적인 프로젝트 폴더 구조
+
+- **tests/**: 코드 테스트 (.R)
+
+```{r tests-folder-files, out.width='70%'}
+knitr::include_graphics("images/testthat-folder-files.png")
+```
+
+---
+
+# 체계적인 프로젝트 폴더 구조
+
+.pull-left[
 
@@
 
-  + **doc/**: 배포용 데이터 분석 문서 (.html)
-
-  + **tests/**: 단위 테스트 (.R)
-
----
-
-# 코드 테스트
-
-.pull-left[
-- 리팩토링(refactoring) 생산성 증가
-
-- R 버전 및 패키지 업데이트에 따른 이상 유무 확인
-
-- One is better than nothing
-]
-
-.pull-right[
-
-```{r refactoring-book, echo = FALSE, fig.cap="Refactoring: improving the design of existing code / Addison-Wesley Professional / Martin Fowler", out.width='40%', fig.align='center'}
+  + **tests/**: 코드 테스트 (.R)
+
+]
+
+.pull-right[
+
+```{r tidy-photo, out.width="90%"}
+knitr::include_graphics("images/mourizal-zativa-gNMVpAPe3PE-unsplash.jpg")
+```
+
+]
+
+---
+class: center, middle
+
+# 리팩토링
+
+---
+
+# 리팩토링
+
+```{r refactoring-book, echo = FALSE, fig.cap="Refactoring: improving the design of existing code / Addison-Wesley Professional / Martin Fowler", out.width='30%', fig.align='center'}
 knitr::include_graphics("images/refact2.jpeg")
@@
 
-]
-
----
-layout: true
-
-# 코드 모호성 제거
-
----
-- 네임스페이스 명시화 / `package::function()`
-
-  + 패키지 로드 순서에 따라 결과가 달라지거나 오류 발생하는 문제
-
----
-- 네임스페이스 명시화 / `package::function()`
-
-  + 어떤 패키지의 함수를 사용할 것인지 명시화
-
---
-
-```{r namespace-3, echo=TRUE, error=TRUE}
-dplyr::tibble(x = 1:2, y = 2:3) %>% 
-  dplyr::select(x)
-```
-
----
-- 네임스페이스 명시화 / `package::function()`
-
-  + 코드를 이해하기 쉽도록 도움
-
---
+---
+
+# 리팩토링
+
+- 코드 테스트
+
+```{r test-results, out.width='50%'}
+knitr::include_graphics("images/test-results.png")
+```
+
+---
+
+# 리팩토링
+
+- 네임 스페이스 명시화
 
@@
 
---
-
 .pull-right[
@@
 ---
-layout: false
-
-# 코드의 재사용성 향상
+
+# 리팩토링
 

### 2021-10-10 draft.Rmd
리팩토링 내용 수정

@@
 library(dmtr)
-library(corrr)
+library(corrr) #<<
 library(dplyr)
@@
   select(x1, x2) %>% 
-  correlate()
+  correlate() #<<
 ```
@@
   select(x1, x2) %>% 
-  corrr::correlate()
+  corrr::correlate() #<<
 ```
@@
 
-  + **tidy eval** 프레임워크 활용
-
-```r
-fisher_ld <- function(.data, .group_var, .xvar) {
-* .group_var <- rlang::enquo(.group_var)
-* .xvar <- rlang::enquo(.xvar)
-* mu_hat <- group_mean(.data, !!.group_var, !!.xvar)
-
-  if (length(mu_hat) != 2L) {
-    stop("number of levels of .group_var must be 2.")
-  }
-
-* sigma_hat <- pooled_variance(.data, !!.group_var, !!.xvar)
-
-  res <- solve(sigma_hat) %*% purrr::reduce(mu_hat, `-`) %>% drop()
-  attr(res, "group") <- attr(mu_hat, "group")
-  return(res)
-}
-
-* fisher_ld(binaryclass2, class, c(x1, x2))
+```{r rlang-logo, out.width='30%'}
+knitr::include_graphics(
+  "images/rlang-logo.png"
+)
 ```

### 2021-10-10 draft.Rmd
맺음말 구성 수정

@@
 ---
+class: center, middle
 
@@
 
-- 분석 과정 및 결과의 배포
+---
+
+# 데이터 분석가에게 패키지 개발이란?
+
+- 분석 과정과 및 결과의 배포
 
@@
 
+--
+
+.pull-left[
+
+```{r cook, fig.alt='https://unsplash.com/photos/gPpbFaEkl00', out.width='80%'}
+knitr::include_graphics("images/nikolay-smeh-gPpbFaEkl00-unsplash.jpg")
+```
+
+]
+
+--
+
+.pull-right[
+
+```{r recipe, fig.alt='https://www.10000recipe.com/recipe/6839413', out.width='100%'}
+knitr::include_graphics("images/bibimbap-recipe-10000recipe.png")
+```
+
+]
+
+---
+
+# 데이터 분석가에게 패키지 개발이란?
+
 - 생산성 향상을 위한 배움의 과정
@@
 
-  + 코드 테스트
-
-  + 코딩 스타일
+  + 리팩토링
+
+--
+
+.pull-left[
+
+```{r messy-photo-2, fig.alt='https://unsplash.com/photos/kn-UmDZQDjM', out.width="90%"}
+knitr::include_graphics("images/xavi-cabrera-kn-UmDZQDjM-unsplash.jpg")
+```
+
+]
+
+--
+
+.pull-right[
+
+```{r tidy-photo-2, fig.alt='https://unsplash.com/photos/gNMVpAPe3PE', out.width="90%"}
+knitr::include_graphics("images/mourizal-zativa-gNMVpAPe3PE-unsplash.jpg")
+```
+
+]
+
+---
+class: center, middle
+
+# 어떻게 시작할까?
+
+---
+
+# 어떻게 시작할까?
+
+--
+
+- 기존에 수행한 간단한 분석 프로젝트로 시작
+
+--
+
+- 현실적인 작업범위 및 기대수준 설정
+
+  + 예) 참고교재, 실무용 아님, 제한된 기능
+
+--
 
@@
 
-  + 쉬운 목표 설정
-
-  + 기존에 수행한 가장 간단한 분석 프로젝트로 시작
-
-  + 하나의 함수부터 시작
-
----
-
-# 현실적인 작업범위 및 기대수준 정의
-
-- 참고교재
-
-- 실무용 아님
-
-- 제한된 유연성
-
----
-
-# 편리한 작업환경 선택
-
-.pull-left[
-
-- RStudio IDE
-
-- `{devtools}`, `{usethis}`, `{testthat}`, etc.
-
-- Git and GitHub
-
-]
+  + 하나의 데이터, 하나의 함수, 하나의 테스트부터 시작
+
+  + 경험자로부터의 피드백
+
+---

### 2021-10-14 draft.Rmd
맺음말 마지막 부분 수정
PDF 버전 추가

@@
 
-```{r data-folder-files, out.width='50%'}
+```{r data-folder-files, out.width='70%'}
 knitr::include_graphics("images/data-folder-files.png")
@@
 
+- 마음가짐: 위축되거나 포기하지 않기
+
+  + 단발성 분석 업무보다 많은 시간과 노력이 소요
+
+  + 경험의 축적이 중요
+
+  + 자신에게 맞는 기대 수준과 페이스 찾기
+
+--
+
 - 기존에 수행한 간단한 분석 프로젝트로 시작
 
---
-
-- 현실적인 작업범위 및 기대수준 설정
-
-  + 예) 참고교재, 실무용 아님, 제한된 기능
-
---
-
-- 경험의 축적이 중요
-
   + 하나의 데이터, 하나의 함수, 하나의 테스트부터 시작
 
-  + 경험자로부터의 피드백
-
----
+  + 사용자부터 피드백 얻기
+
+--
+
+- 분석 업무와 패키지 개발의 선순환
+
+  + 습득한 개발 방법을 분석 업무에 적용
+
+  + 분석 업무의 유연성 확장
+
+  + 새로운 분석 업무에서 패키지 개발의 아이디어 얻기
+
+---

### 2021-10-28 draft.Rmd
마무리 수정

@@
 
-데이터 생성
-{{content}}
-
---
-
-**함수 1**
-{{content}}
-
---
-
-**함수 2**
-{{content}}
-
---
-
-분석
-{{content}}
-
---
-
-**함수 3**
-{{content}}
-
---
-
-분석
-{{content}}
-
---
-
-데이터 생성
-{{content}}
-
---
-
-**함수 4**
-{{content}}
-
---
-
-**함수 5**
-{{content}}
-
---
-
-데이터 분석
+`df1 <- tibble()` \# 데이터 생성
+{{content}}
+
+--
+
+`f1 <- function() {}` \# **함수 정의**
+{{content}}
+
+--
+
+`f2 <- function() {}` \# **함수 정의**
+{{content}}
+
+--
+
+`df1 %>% f1() %>% f2()` \# 분석
+{{content}}
+
+--
+
+`f3 <- function() {}` \# **함수 정의**
+{{content}}
+
+--
+
+`df1 %>% f1() %>% f3()` \# 분석
+{{content}}
+
+--
+
+`df2 <- tibble()` \# 데이터 생성
+{{content}}
+
+--
+
+`f4 <- function() {}` \# **함수 정의**
+{{content}}
+
+--
+
+`f5 <- function() {}` \# **함수 정의**
+{{content}}
+
+--
+
+`df2 %>% f1() %>% f4() %>% f5()` \# 분석
 {{content}}
@@
 ```{r}`r ''`
-library(dmtr)
+*library(dmtr)
 
@@
   local_edition(3)
-  expect_equal(
+  expect_equal( #<<
     fisher_ld(binaryclass2, class, x1:x2), #<<
@@
 
-```{r recipe, fig.alt='https://www.10000recipe.com/recipe/6839413', out.width='100%'}
+```{r recipe, fig.alt='https://www.10000recipe.com/recipe/6839413', out.width='100%', fig.cap='https://www.10000recipe.com/recipe/6839413'}
 knitr::include_graphics("images/bibimbap-recipe-10000recipe.png")
@@
 ---
+class: center, middle, inverse
+
+# 감사합니다

### 2021-11-03 draft.Rmd
발표 흐름에 따라 슬라이드 내용 수정

@@
 
-```{r lda-example-plot, echo=FALSE, out.width='80%', fig.align='center', fig.cap="예제: 선형 분류분석"}
-data(binaryclass2, package = "dmtr")
-
-binaryclass2 %>% 
-  ggplot(aes(x = x1, y = x2, color = class)) +
-  geom_abline(slope = 0.9781, intercept = 0.3414, 
-              color = "grey20", linetype = "dashed") +
-  geom_point(size = 5) +
-  geom_text(aes(label = class), vjust = -1, size = 5) +
-  scale_x_continuous(
-    breaks = seq(1, 10, by = 1),
-    minor_breaks = NULL,
-    expand = expansion(mult = 0.05)
-  ) +
-  scale_y_continuous(
-    breaks = seq(1, 10, by = 1),
-    minor_breaks = NULL,
-    expand = expansion(mult = 0.05)
-  ) +
-  scale_color_manual(
-    values = c(`1` = "steelblue", `2` = "firebrick")
-  ) +
-  theme_light() +
-  theme(
-    legend.position = "none",
-    text = element_text(size = 20),
-    aspect.ratio = 1
-  )
-```
-
-]
-
----
-
-.pull-right[
-
 - 데이터 마이닝 이론과 알고리즘을 잘 이해할 수 있게 깔끔하게 구성된 교재
@@
 
-library(MASS)
-fit <- lda(class ~ x1 + x2, binaryclass2)
+library(MASS) #<<
+fit <- lda(class ~ x1 + x2, binaryclass2) #<<
 
@@
 )
-* df %>% tidyr::unnest(a, b)
+*df %>% tidyr::unnest(a, b)
 ```
@@
 ---
-
-# 지금은 맞지만, 나중엔 틀릴수도?
-
-- 오류 발생 시 수정에 많은 시간 소요
-
-- 코드 실행 오류는 없으나 결과가 틀린 경우는 오류 발견 자체가 어려움
+class: center, middle
+
+# 해결방안: 패키지 개발
 
@@
 
-```{r data-raw-example, out.width='30%', fig.align='center'}
+```{r data-raw-example, out.width='20%', fig.align='center'}
 knitr::include_graphics(c(
@@
 
-```{r r-example, out.width='30%', fig.align='center'}
+```{r r-example, out.width='20%', fig.align='center'}
 knitr::include_graphics(c(
@@
 
-```{r vignettes-example, out.width='30%', fig.align='center'}
+```{r vignettes-example, out.width='20%', fig.align='center'}
 knitr::include_graphics(c(
@@
 
-```{r vignettes-example-2, out.width='30%', fig.align='center'}
+```{r vignettes-example-2, out.width='20%', fig.align='center'}
 knitr::include_graphics(c(
@@
 
-```{r man-example, out.width='30%', fig.align='center'}
+```{r man-example, out.width='20%', fig.align='center'}
 knitr::include_graphics(c(
@@
 ---
-
----
-
-```{r tests-example, out.width='10%', fig.align='center'}
-knitr::include_graphics(c(
-  "images/tests-folder-icon.png", 
-  "images/tests-discriminant-functions.png"))
-```
 
@@
 ---
+
+```{r tests-example, out.width='8%', fig.align='center'}
+knitr::include_graphics(c(
+  "images/tests-folder-icon.png", 
+  "images/tests-discriminant-functions.png"))
+```
+
+---
 class: middle, center
@@
 
-- 분석 과정과 및 결과의 배포
+- 분석 과정과 결과의 배포
 
@@
 
-  + 사용자부터 피드백 얻기
-
 --
@@
 
-  + 분석 업무의 유연성 확장
-
-  + 새로운 분석 업무에서 패키지 개발의 아이디어 얻기
+  + 분석 업무에서 패키지 아이디어 확장
 
@@
 # 감사합니다
+
+# **
+
+### Spectial thanks to 
+### 전치혁 교수님
+### 이혜선 교수님
+### 이종석 교수님

### 2021-11-03 draft.Rmd
애니메이션 제거 - 발표시간 단축을 위해

@@
 layout: false
-
-# 데이터마이닝 교재 참고자료 제작
-
-.pull-left[
-
-```{r rlogo, out.width="70%"}
-knitr::include_graphics(c("images/Rlogo.png"))
-```
-
-]
-
-.pull-right[
-
-```{r bookdownlogo, out.width="70%"}
-knitr::include_graphics(c("images/bookdown-logo.png"))
-```
-
-]
-
----
 # 데이터마이닝 교재 참고자료 제작
@@
 .pull-right[
+
+```{r rlogo-bookdown, out.width="30%"}
+knitr::include_graphics(c("images/Rlogo.png", "images/bookdown-logo.png"))
+```
+
   {{content}}
@@
 
-{{content}}
-
-]
-
---
-
 `df1 <- tibble()` \# 데이터 생성
-{{content}}
-
---
 
 `f1 <- function() {}` \# **함수 정의**
-{{content}}
-
---
 
 `f2 <- function() {}` \# **함수 정의**
-{{content}}
-
---
 
 `df1 %>% f1() %>% f2()` \# 분석
-{{content}}
-
---
 
 `f3 <- function() {}` \# **함수 정의**
-{{content}}
-
---
 
 `df1 %>% f1() %>% f3()` \# 분석
-{{content}}
-
---
 
 `df2 <- tibble()` \# 데이터 생성
-{{content}}
-
---
 
 `f4 <- function() {}` \# **함수 정의**
-{{content}}
-
---
 
 `f5 <- function() {}` \# **함수 정의**
-{{content}}
-
---
 
 `df2 %>% f1() %>% f4() %>% f5()` \# 분석
-{{content}}
-
---
 
 **...**
-{{content}}
+
+]
 

### 2021-11-03 draft.Rmd
수정 - 발표시간 단축 위해 "체계적인 프로젝트 폴더 구조" 부분 간소화

@@
 
-- 프로젝트 폴더
-
-```{r package-folder-structure, out.width='70%'}
-knitr::include_graphics("images/package-folder-structure.png")
-```
-
----
-
-# 체계적인 프로젝트 폴더 구조
-
-- **data-raw/**: 데이터 생성/쿼리 R 스크립트 (.R)
+.pull-left[
+
+- 일관된 프로젝트 폴더 구조
+
+  + **data-raw/: 데이터 생성/쿼리 R 스크립트 (.R)**
+
+]
+
+.pull-right[
 
@@
 
+]
+
 ---
@@
 
-- **data/**: 생성된 데이터 파일 (.rda)
+.pull-left[
+
+- 일관된 프로젝트 폴더 구조
+
+  + data-raw/: 데이터 생성/쿼리 R 스크립트 (.R)
+
+  + **data/: 생성된 데이터 파일 (.rda)**
+
+]
+
+.pull-right[
 
@@
 
-- **R/**: 재사용을 위해 모듈화된 데이터 분석 함수 (.R)
+.pull-left[
+
+- 일관된 프로젝트 폴더 구조
+
+  + data-raw/: 데이터 생성/쿼리 R 스크립트 (.R)
+
+  + data/: 생성된 데이터 파일 (.rda)
+
+  + **R/: 재사용을 위해 모듈화된 데이터 분석 함수 (.R)**
+
+]
+
+.pull-right[
 
@@
 
-- **vignettes/**: 데이터 분석 문서 소스 (.Rmd)
+.pull-left[
+
+- 일관된 프로젝트 폴더 구조
+
+  + data-raw/: 데이터 생성/쿼리 R 스크립트 (.R)
+
+  + data/: 생성된 데이터 파일 (.rda)
+
+  + R/: 재사용을 위해 모듈화된 데이터 분석 함수 (.R)
+
+  + **vignettes/: 데이터 분석 문서 소스 (.Rmd)**
+
+]
+
+.pull-right[
 
@@
 
-- **tests/**: 코드 테스트 (.R)
+.pull-left[
+
+- 일관된 프로젝트 폴더 구조
+
+  + data-raw/: 데이터 생성/쿼리 R 스크립트 (.R)
+
+  + data/: 생성된 데이터 파일 (.rda)
+
+  + R/: 재사용을 위해 모듈화된 데이터 분석 함수 (.R)
+
+  + vignettes/: 데이터 분석 문서 소스 (.Rmd)
+
+  + **tests/: 코드 테스트 (.R)**
+
+]
+
+.pull-right[
 
@@
 ```
+
+]
 

### 2021-11-04 draft.Rmd
Revert "rename file: draft.Rmd -> index.Rmd"
This reverts commit 181d1eb4af45be2fa7f940ca09922826c6002841.

revert commit, to keep draft.Rmd history

@@
 ---
-layout: true
-
 # 데이터마이닝 교재 참고자료 제작
@@
 
----
-
----
+--
 
@@
 ---
-layout: false
 # 데이터마이닝 교재 참고자료 제작
@@
 
-# 사용자 함수의 난립
-
----
-# 사용자 함수의 난립
+# 정돈되지 않은 코드
+
+---
+# 정돈되지 않은 코드
 
@@
 
-# 사용자 함수의 난립
+# 정돈되지 않은 코드
 
@@
 
-```{r r-function-manual, echo = FALSE, fig.cap="함수 도움말 작성 및 컴파일된 도움말 파일 (.Rd)", out.width='90%', fig.align='center'}
+```{r r-function-manual, echo = FALSE, fig.cap="컴파일된 도움말 파일 (.Rd)", out.width='90%', fig.align='center'}
 knitr::include_graphics("images/r-function-manual-fisher_ld.png")
@@
 
-# 코드 테스트
-
----
-
-# 코드 테스트
+# 코드 테스트 작성
+
+---
+
+# 코드 테스트 작성
 
@@
 
-# 코드 테스트
+# 코드 테스트 작성
 
@@
 
---
-
 - 마음가짐: 위축되거나 포기하지 않기
@@
 
-### Spectial thanks to 
-### 전치혁 교수님
-### 이혜선 교수님
-### 이종석 교수님
+#### 특히 고마운 분들: 
+#### 전치혁 교수님
+#### 이혜선 교수님
+#### 이종석 교수님

### 2021-11-06 index.Rmd
애니메이션 일부 삭제 - 발표 시간 축소 위함

--- 2021-11-04 draft.Rmd
@@
 - 데이터 마이닝 이론과 알고리즘을 잘 이해할 수 있게 깔끔하게 구성된 교재
-{{content}}
-
-]
-
---
 
 - 간단하면서 결과를 눈으로 확인할 수 있는 예제들
-{{content}}
-
---
 
 - **예제 결과 도출을 위한 프로그램 구현은 제공되지 않음**
-{{content}}
+
+]
 
@@
 
-  {{content}}
-]
-
---
-
 - 온라인 버전 보조 교재로 활용
-{{content}}
-
---
 
 - 예제를 R로 구현한 내용을 추가
-{{content}}
-
---
 
 - 단순히 기존 R 패키지를 사용하는 것보다 깊이 있는 내용을 제공
-{{content}}
+
+]
 
